---
title: "SCCS Data"
author: |
  Hesam Mahmoudi  
  Lung Cancer Policy Team (LCP)  
  MGB Center for Health Technology Assessment (CHTA)  
  Massachusetts General Hospital (MGH),  
  Harvard Medical School  
date: "Started: 2024-12-14 \\ Finalized: 2025-05-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load core libraries
library(tidyverse)     # includes ggplot2, dplyr, tidyr, tibble, stringr, etc.
library(patchwork)     # for combining plots
library(kableExtra)    # for styled tables
library(flextable)     # for Word table formatting
library(officer)       # for Word/PowerPoint export
library(openxlsx)      # for Excel export
library(haven)         # for reading SPSS/SAS/Stata files
library(htmltools)     # For building custom HTML widgets
library(webshot2)
library(cowplot)
library(gtable)
library(grid)
library(ggtext)
```


## Functions
#SHG_Setup
This function prepares the input file (input_r.txt) for the simulation executable. It accepts parameters like Sex, YOB, Population, and optional values such as Seed, Race, Path, and Version. The function modifies a template file, updates it with the specified values, and writes the customized file to the simulation directory.
 
```{r}
SHG_Setup <- function(Sex, YOB, Population, 
                      Seed = 1, Race = 0, Path = "SHG6.3.4", Version = 2023) {
  
  # Validate inputs
  if (!Version %in% c(2023, 2017)) {
    stop("Version must be 2023 or 2017.")
  }
  
  # Determine the file paths based on the Version
  prob_paths <- if (Version == 2023) {
    list(
      INIT_PROB = "SHG6.3.4/data/06_21_2023/lbc_shg_initiation06212023.txt",
      CESS_PROB = "SHG6.3.4/data/06_21_2023/lbc_shg_cessation06212023.txt",
      OCD_PROB  = "SHG6.3.4/data/06_21_2023/lbc_smokehist_ac_mortality_02212016.txt",
      CPD_DATA  = "SHG6.3.4/data/06_21_2023/lbc_shg_cpd06212023.txt"
    )
  } else {
    list(
      INIT_PROB = "SHG6.3.4/data/05_03_2017/lbc_shg_initiation05032017.txt",
      CESS_PROB = "SHG6.3.4/data/05_03_2017/lbc_shg_cessation05032017.txt",
      OCD_PROB  = "SHG6.3.4/data/05_03_2017/lbc_smokehist_ac_mortality_02212016.txt",
      CPD_DATA  = "SHG6.3.4/data/05_03_2017/lbc_shg_cpd05032017.txt"
    )
  }
  
  # Path to the input template
  input_template_path <- file.path(Path, "SHG_input_temp.txt")
  if (!file.exists(input_template_path)) {
    stop(paste("Template file not found at:", input_template_path))
  }
  
  # Read the input template
  input_text <- readLines(input_template_path)
  
  # Replace placeholders in the template
  modified_text <- input_text %>%
    gsub("{seed1}", Seed, ., fixed = TRUE) %>%
    gsub("{seed2}", Seed, ., fixed = TRUE) %>%
    gsub("{seed3}", Seed, ., fixed = TRUE) %>%
    gsub("{seed4}", Seed, ., fixed = TRUE) %>%
    gsub("RACE=0", paste0("RACE=", Race), ., fixed = TRUE) %>%
    gsub("SEX=0", paste0("SEX=", Sex), ., fixed = TRUE) %>%
    gsub("{cohort}", YOB, ., fixed = TRUE) %>%
    gsub("REPEAT=100000", paste0("REPEAT=", Population), ., fixed = TRUE) %>%
    gsub("INIT_PROB=SHG6.3.4/data/06_21_2023/lbc_shg_initiation06212023.txt", 
         paste0("INIT_PROB=", prob_paths$INIT_PROB), ., fixed = TRUE) %>%
    gsub("CESS_PROB=SHG6.3.4/data/06_21_2023/lbc_shg_cessation06212023.txt", 
         paste0("CESS_PROB=", prob_paths$CESS_PROB), ., fixed = TRUE) %>%
    gsub("OCD_PROB=SHG6.3.4/data/06_21_2023/lbc_smokehist_oc_mortality_02212016.txt", 
         paste0("OCD_PROB=", prob_paths$OCD_PROB), ., fixed = TRUE) %>%
    gsub("CPD_DATA=SHG6.3.4/data/06_21_2023/lbc_shg_cpd06212023.txt", 
         paste0("CPD_DATA=", prob_paths$CPD_DATA), ., fixed = TRUE)
  
  # Path to the output file
  output_file_path <- file.path(Path, "input_r.txt")
  
  # Write the modified text to the output file
  writeLines(modified_text, output_file_path)
  
  # Notify the user
  message("Modified input file written to: ", output_file_path)
}
```


#SHG_Run
This function executes the simulation executable (lbc_smokehist_win.exe) using a specified input file (input_r.txt). After running the simulation, it checks for the existence of the output file (output_r.out) to confirm success. The function reports an error if the output file is not generated.

```{r}
SHG_Run <- function(Path = "SHG6.3.4") {
  # Define paths for the input file, executable, and output file
  input_file <- file.path(Path, "input_r.txt")
  exe_file <- file.path(Path, "lbc_smokehist_win.exe")
  output_file <- file.path("output_r.out")
  
  # Check if the executable and input file exist
  if (!file.exists(exe_file)) {
    stop(paste("Executable not found at:", exe_file))
  }
  if (!file.exists(input_file)) {
    stop(paste("Input file not found at:", input_file))
  }
  
  # Build the command to execute
  command <- paste(shQuote(exe_file), shQuote(input_file))
  
  # Run the command
  message("Running command: ", command)
  system(command, wait = TRUE)
  
  # Check for the output file
  if (file.exists(output_file)) {
    message("Success: Output file created at ", output_file)
  } else {
    stop("Error: Output file not found after running the executable.")
  }
}
```


#SHG_Output
This function reads and processes the output file (output_r.out) generated by the simulation executable. It extracts individual-level simulation results such as smoking initiation, cessation, and death ages, as well as yearly smoking data (cigarettes per day). The results are returned in a structured dataframe for further analysis.

```{r}
SHG_Output <- function(outfile = "output_r.out") {
  # Check if the output file exists
  if (!file.exists(outfile)) {
    stop(paste("Output file not found:", outfile))
  }
  
  # Read the contents of the output file
  file_lines <- readLines(outfile)
  
  # Find the <RUN> section
  run_start <- grep("<RUN>", file_lines)
  run_end <- grep("</RUN>", file_lines)
  
  if (length(run_start) == 0 || length(run_end) == 0) {
    stop("The <RUN> section is missing in the output file.")
  }
  
  # Extract the lines within the <RUN> section
  run_lines <- file_lines[(run_start + 1):(run_end - 1)]
  
  # Initialize an empty list to hold the data for each individual
  rows <- list()
  
  # Loop through each line in the <RUN> section
  for (line in run_lines) {
    # Split the line by semicolons
    parts <- unlist(strsplit(line, ";"))
    
    # Extract the first six values
    race <- as.numeric(parts[1])
    sex <- as.numeric(parts[2])
    yob <- as.numeric(parts[3])
    initiation_age <- as.numeric(parts[4])
    cessation_age <- as.numeric(parts[5])
    death_age <- as.numeric(parts[6])
    if (death_age==-999){
      death_age <- 109
    }
    
    # Initialize a vector to store CPD values for ages 1 to 110
    cpd_values <- rep(0, 110)
    
    # Check if there are CPD values in the line
    if (length(parts) > 6) {
      # Extract the remaining values after the first six
      age_cpd_pairs <- as.numeric(parts[7:length(parts)])
      
      # Loop through the age-CPD pairs
      for (i in seq(1, length(age_cpd_pairs), by = 2)) {
        age <- age_cpd_pairs[i]
        cpd <- age_cpd_pairs[i + 1]
        
        # Stop if we've passed cessation or death age
        if (!is.na(cessation_age) && age > cessation_age && cessation_age!=-999) break
        if (!is.na(death_age) && age > death_age) break
        
        # Assign the CPD value for the given age
        cpd_values[age] <- cpd
      }
    }
    
    # Combine the data into a row
    row <- c(race, sex, yob, initiation_age, cessation_age, death_age, cpd_values)
    rows <- append(rows, list(row))
  }
  
  # Create a dataframe from the rows
  column_names <- c("Race", "Sex", "YOB", "Initiation_Age", "Cessation_Age", "Death_Age", 1:110)
  SHG_out <- as.data.frame(do.call(rbind, rows))
  colnames(SHG_out) <- column_names
  
  return(SHG_out)
}
```


#SHG_rslts_ini
This function initializes an empty dataframe for storing simulation results. The dataframe contains columns for key variables such as Sex, Race, YOB, Age, Initiation_Age, Cessation_Age, Eversmoker, Smokenow, CigsPerDay, Pack_Years, and Smoking_Years. It provides a structured container to accumulate and manage simulation data.

```{r}
SHG_rslts_ini <- function() {
    # Initialize aggr_out with specified columns
    df_name <- data.frame(
      Sex = integer(),
      Race = integer(),
      YOB = integer(),
      Age = integer(),
      Initiation_Age = integer(),
      Cessation_Age = integer(),
      Quit_Age = integer(),
      Eversmoker = integer(),
      Smokenow = integer(),
      CigsPerDay = numeric(),
      Pack_Years_Minusone = numeric(),
      Pack_Years = numeric(),
      Smoking_Years = integer(),
      Smoking_Years_Minusone = integer(),
      stringsAsFactors = FALSE
    )
    return(df_name)
  }
```


#SHG_Aggr
This function aggregates simulation results (shgrun_out) into a dataframe (aggr_out) for analysis. It calculates individual attributes like smoking status, cigarettes per day, pack years, and smoking years based on simulation inputs and outputs. The function also ensures that only individuals alive at the data collection year are included. If aggr_out is not already initialized, it creates it.

```{r}
SHG_Aggr <- function(shgrun_out, aggr_out, collection_year, exclude_deceased = "On") {
  for (i in seq_len(nrow(shgrun_out))) {
    # Extract values from shgrun_out
    row <- shgrun_out[i, ]
    sex <- row$Sex
    race <- row$Race
    yob <- row$YOB
    initiation_age <- row$Initiation_Age
    cessation_age <- row$Cessation_Age
    death_age <- row$Death_Age

    # Calculate age
    age <- collection_year - yob

    # Exclude deceased individuals if the option is "On"
    if (exclude_deceased == "On" && age > death_age) {
      next  # Skip this individual
    }

    # Determine if the individual is an EverSmoker
    eversmoker <- ifelse(initiation_age == -999 | initiation_age > age, 0, 1)

    # Initialize derived values
    smokenow <- 0
    cigs_per_day <- 0
    pack_years <- 0
    pack_years_minusone <- 0
    smoking_years <- 0
    smoking_years_minusone <- 0
    quit_age <- 0
    
    # Check if the individual has smoked all the way through their life
    if (cessation_age < 0 & initiation_age > 0) {
      cessation_age <- death_age
    }
    
    if (eversmoker == 1) {
      # Check if the individual is currently smoking
      if (age > initiation_age & age <= cessation_age) {
        smokenow <- 1
      }
      
      quit_age <- cessation_age+1

      # Extract smoking data from ages 1 to current age
      smoking_data <- as.numeric(row[7:(6 + age)])  # Columns 7 to (6 + age)

      # Calculate non-zero average for CigsPerDay
      cigs_per_day <- mean(smoking_data[smoking_data > 0], na.rm = TRUE)

      # Calculate Pack Years (sum of all smoking data divided by 20)
      pack_years <- sum(smoking_data, na.rm = TRUE) / 20

      # Calculate Smoking Years
      if (!is.na(cessation_age)) {
        if (cessation_age >= age) {
          smoking_years <- age - initiation_age + 1  # Individual hasn't quit smoking yet
          smoking_years_minusone <- age - initiation_age
        } else if (cessation_age > initiation_age) {
          smoking_years <- cessation_age - initiation_age + 1 # Normal case
          smoking_years_minusone <- cessation_age - initiation_age
        }
      }
      
      # Calculate Pack Years (smoking years * CPD/20)
      pack_years_minusone <- smoking_years_minusone * cigs_per_day / 20
    }

    # Append the individual's data to aggr_out
    aggr_out <- rbind(aggr_out, data.frame(
      Sex = sex,
      Race = race,
      YOB = yob,
      Age = age,
      Initiation_Age = initiation_age,
      Cessation_Age = cessation_age,
      Quit_Age = quit_age,
      Eversmoker = eversmoker,
      Smokenow = smokenow,
      CigsPerDay = cigs_per_day,
      Pack_Years = pack_years,
      Pack_Years_Minusone = pack_years_minusone,
      Smoking_Years = smoking_years,
      Smoking_Years_Minusone = smoking_years_minusone,
      stringsAsFactors = FALSE
    ))
  }
  return(aggr_out)
}
```


#SHGdat2BY
This function creates a birth year report from simulation results. It takes a dataframe containing simulation results (shgdata) and groups the data by YOB (Year of Birth). For each year of birth, it calculates the count of Females and Males and returns a summarized dataframe ordered by YOB.

```{r}
SHGdat2BY <- function(shgdata) {
  # Check if required columns are present
  required_columns <- c("YOB", "Sex")
  missing_columns <- setdiff(required_columns, colnames(shgdata))
  if (length(missing_columns) > 0) {
    stop(paste("The following required columns are missing from 'shgdata':", paste(missing_columns, collapse = ", ")))
  }
  
  # Check if YOB and Sex columns contain valid data types
  if (!is.numeric(shgdata$YOB)) {
    stop("Column 'YOB' must be numeric.")
  }
  if (!is.numeric(shgdata$Sex)) {
    stop("Column 'Sex' must be numeric.")
  }
  
  # Ensure no negative or nonsensical years in YOB
  if (any(shgdata$YOB < 1800 | shgdata$YOB > 2100, na.rm = TRUE)) {
    warning("Some 'YOB' values are outside a reasonable range (1800-2100). These will be included unless removed beforehand.")
  }
  
  # Summarize data by YOB
  SHGBY <- shgdata %>%
    group_by(YOB) %>%
    summarise(
      Females = sum(Sex == 1, na.rm = TRUE),  # Count of females
      Males = sum(Sex == 0, na.rm = TRUE)    # Count of males
    ) %>%
    arrange(YOB)  # Sort by YOB
  
  # Return the summarized dataframe
  return(SHGBY)
}
```


#sufftrgt
This function scales the values in a birth year report (byrdata) to create a target dataset. It multiplies the counts of Females and Males in the input dataframe by a coverage factor (default is 10) and returns the scaled dataset.

```{r}
sufftrgt <- function(byrdata, coverage = 10) {
  # Check for required columns
  required_columns <- c("Females", "Males")
  missing_columns <- setdiff(required_columns, colnames(byrdata))
  if (length(missing_columns) > 0) {
    stop(paste("The following required columns are missing from 'byrdata':", paste(missing_columns, collapse = ", ")))
  }

  # Scale Females and Males by the coverage factor
  byrtrgt <- byrdata
  byrtrgt$Females <- coverage * byrdata$Females
  byrtrgt$Males <- coverage * byrdata$Males
  
  # Return the modified dataframe
  return(byrtrgt)
}
```


#suff_shortage
This function calculates the shortage of individuals in a birth year dataset relative to a target dataset. It takes two dataframes: byrrslts (results) and byrtrgt (target). For each YOB-gender combination, if the results fall short of the target, it calculates the difference; otherwise, it assigns zero. The output matches the structure of the target dataset (byrtrgt).

```{r}
suff_shortage <- function(byrrslts, byrtrgt) {
  # Check for required columns
  required_columns <- c("YOB", "Females", "Males")
  missing_byrrslts <- setdiff(required_columns, colnames(byrrslts))
  missing_byrtrgt <- setdiff(required_columns, colnames(byrtrgt))
  if (length(missing_byrrslts) > 0) {
    stop(paste("The following required columns are missing from 'byrrslts':", paste(missing_byrrslts, collapse = ", ")))
  }
  if (length(missing_byrtrgt) > 0) {
    stop(paste("The following required columns are missing from 'byrtrgt':", paste(missing_byrtrgt, collapse = ", ")))
  }
  
  # Merge the two dataframes by YOB, keeping all rows from byrtrgt
  combined <- merge(
    byrtrgt, byrrslts, 
    by = "YOB", 
    all.x = TRUE,  # Keep all rows from byrtrgt
    suffixes = c("_trgt", "_rslts")
  )
  
  # Replace NA values in byrrslts columns with 0
  combined$Females_rslts[is.na(combined$Females_rslts)] <- 0
  combined$Males_rslts[is.na(combined$Males_rslts)] <- 0
  
  # Calculate shortage
  combined$Females_shortage <- pmax(combined$Females_trgt - combined$Females_rslts, 0)
  combined$Males_shortage <- pmax(combined$Males_trgt - combined$Males_rslts, 0)
  
  # Create the output dataframe with the same structure as byrtrgt
  byrrem <- combined[, c("YOB", "Females_shortage", "Males_shortage")]
  colnames(byrrem) <- c("YOB", "Females", "Males")
  
  return(byrrem)
}
```


#SHGbyr
This function runs simulations to meet population targets specified in a birth year dataset (BYr). For each YOB-gender combination, it prepares input files, runs simulations, processes outputs, and aggregates results into AggData. It uses looping to process all rows and genders in BYr and dynamically updates the aggregated results.

```{r}
SHGbyr <- function(BYr, prcssyr, seed, exclude_deceased = "On") {
  # Input Validation ---------------------------------------------------------
  if (!all(c("YOB") %in% colnames(BYr))) {
    stop("The 'BYr' dataframe must include a 'YOB' column.")
  }
  if (ncol(BYr) < 3) {
    stop("The 'BYr' dataframe must have at least three columns: 'YOB', 'Females', and 'Males'.")
  }

  # Initialize an empty dataframe for aggregated data
  AggData <- SHG_rslts_ini()
  
  # Outer loop: Iterate over each row in the 'BYr' dataframe
  for (i in 1:nrow(BYr)) {
    # Extract the birth year for the current row
    YBirth <- BYr[i, "YOB"]
    
    # Inner loop: Iterate over genders (j = 1 for males, j = 2 for females)
    for (j in c(1, 2)) {
      # Extract the population for the current year and gender
      poprun <- BYr[i, j + 1]
      
      # Skip if the population is zero or NA
      if (is.na(poprun) || poprun == 0) {
        next
      }
      
      # Map gender: j = 1 -> gender = 1 (female), j = 2 -> gender = 0 (male)
      gender <- 2 - j
      
      # Set up the simulation input
      SHG_Setup(Sex = gender, YOB = YBirth, Population = poprun, Seed = i * seed + j)
      
      # Run the simulation
      SHG_Run()
      
      # Process the simulation output
      SHGout <- SHG_Output()
      
      # Aggregate the simulation output
      AggData <- SHG_Aggr(SHGout, AggData, prcssyr, exclude_deceased)  # Pass exclude_deceased
    }
  }
  
  # Return the aggregated data
  return(AggData)
}
```


#SHGbyr_extend
This function extends the aggregated results (AggData) until population shortages specified in BYr are resolved. It calculates shortages using suff_shortage and runs additional simulations for any YOB-gender combination that falls short. The process continues until all shortages are eliminated, and the function returns the updated AggData.

```{r}
SHGbyr_extend <- function(AggData, BYr, prcssyr, seed, exclude_deceased = "On") {
  # Input Validation ---------------------------------------------------------
  if (!all(c("YOB", "Females", "Males") %in% colnames(BYr))) {
    stop("The 'BYr' dataframe must include the columns: 'YOB', 'Females', and 'Males'.")
  }
  
  # Processing Loop ----------------------------------------------------------
  # Step 1: Initialize shortage calculation
  repeat {
    # Step 2: Get the current birth year report from AggData
    Aggbyr <- SHGdat2BY(AggData)
    
    # Step 3: Calculate the shortage
    newbyr <- suff_shortage(Aggbyr, BYr)
    
    # Check sufficiency criteria: All shortages must be zero
    if (all(newbyr$Females == 0 & newbyr$Males == 0)) {
      break  # Exit the loop when no shortages remain
    }
    
    # Step 4: Iterate over each row in 'newbyr'
    for (i in seq_len(nrow(newbyr))) {
      # Extract the birth year for the current row
      YBirth <- newbyr[i, "YOB"]
      
      # Inner loop: Iterate over genders (1 for females, 2 for males)
      for (j in c(1, 2)) {
        # Extract the population for the current year and gender
        poprun <- newbyr[i, j + 1]  # Column 2 is Females, Column 3 is Males
        
        # Skip if the population is zero or NA
        if (is.na(poprun) || poprun == 0) {
          next
        }
        
        # Map gender: 1 -> Female (1), 2 -> Male (0)
        gender <- 2 - j
        
        # Set up the simulation input
        SHG_Setup(Sex = gender, YOB = YBirth, Population = poprun, Seed = i * seed + j)
        
        # Run the simulation
        SHG_Run()
        
        # Process the simulation output
        SHGout <- SHG_Output()
        
        # Aggregate the simulation output, incorporating exclude_deceased
        AggData <- SHG_Aggr(SHGout, AggData, prcssyr, exclude_deceased = exclude_deceased)
      }
    }
    
    # Increment the seed to ensure variability in simulations
    seed <- seed + 1
  }
  
  # Return the aggregated data ------------------------------------------------
  return(AggData)
}
```


#sample_SHG
This function performs random sampling from simulation results (SHG_rslts) based on a target dataset (BYr). For each YOB-gender combination in BYr, it filters eligible candidates from SHG_rslts and selects a random sample matching the target size. The sampled entries are aggregated and returned as a new dataframe.

```{r}
sample_SHG <- function(SHG_rslts, BYr, seed) {
  # Input Validation ---------------------------------------------------------
  if (!is.data.frame(SHG_rslts)) {
    stop("Input 'SHG_rslts' must be a dataframe.")
  }
  if (!all(c("YOB", "Females", "Males") %in% colnames(BYr))) {
    stop("The 'BYr' dataframe must include columns: 'YOB', 'Females', and 'Males'.")
  }
  if (!is.numeric(seed) || length(seed) != 1) {
    stop("Input 'seed' must be a single numeric value.")
  }
  
  # Set the random seed for reproducibility
  set.seed(seed)
  
  # Initialize an empty dataframe to store the output samples
  sampled_data <- SHG_rslts[0, ]  # Empty dataframe with same structure as SHG_rslts
  
  # Sampling Loop ------------------------------------------------------------
  for (i in 1:nrow(BYr)) {
    # Extract the current YOB and sample populations
    current_yob <- as.numeric(BYr[i, "YOB"])
    female_pop <- as.numeric(BYr[i, "Females"])
    male_pop <- as.numeric(BYr[i, "Males"])
    
    # Sample females
    if (female_pop > 0) {
      female_candidates <- SHG_rslts[which(SHG_rslts$YOB == current_yob & SHG_rslts$Sex == 1), ]
      if (nrow(female_candidates) == 0) {
        warning(paste("No female candidates for YOB =", current_yob))
      } else {
        female_pop <- min(female_pop, nrow(female_candidates))  # Adjust to available candidates
        female_sample <- female_candidates[sample(seq_len(nrow(female_candidates)), size = female_pop), ]
        sampled_data <- rbind(sampled_data, female_sample)
      }
    }
    
    # Sample males
    if (male_pop > 0) {
      male_candidates <- SHG_rslts[which(SHG_rslts$YOB == current_yob & SHG_rslts$Sex == 0), ]
      if (nrow(male_candidates) == 0) {
        warning(paste("No male candidates for YOB =", current_yob))
      } else {
        male_pop <- min(male_pop, nrow(male_candidates))  # Adjust to available candidates
        male_sample <- male_candidates[sample(seq_len(nrow(male_candidates)), size = male_pop), ]
        sampled_data <- rbind(sampled_data, male_sample)
      }
    }
  }
  
  # Return the sampled data --------------------------------------------------
  return(sampled_data)
}
```


#smpl2age
This function summarizes smoking-related metrics by Age and Sex for a given simulation sample or results dataframe. It calculates the percentage of Eversmoker and Smokenow individuals, as well as the average CigsPerDay, Pack_Years, and Smoking_Years for those who are Eversmoker == 1. The resulting dataframe provides an age-sex breakdown of smoking behaviors and history.

```{r}
smpl2age <- function(data) {
  # Input Validation ---------------------------------------------------------
  if (!is.data.frame(data)) {
    stop("Input 'data' must be a dataframe.")
  }
  
  required_columns <- c("Age", "Sex", "Eversmoker", "Smokenow", 
                        "CigsPerDay", "Pack_Years", "Smoking_Years")
  missing_columns <- setdiff(required_columns, colnames(data))
  
  if (length(missing_columns) > 0) {
    stop(paste("Missing required columns:", paste(missing_columns, collapse = ", ")))
  }
  
  # Create a summary dataframe -----------------------------------------------
  summary_df <- data %>%
    group_by(Age, Sex) %>% 
    summarise(
      Total = n(),
      EverSmoker_Percentage = 100 * mean(Eversmoker == 1, na.rm = TRUE),
      SmokeNow_Percentage = 100 * mean(Smokenow == 1, na.rm = TRUE),
      CPD_Avg = mean(CigsPerDay[Eversmoker == 1], na.rm = TRUE),
      CPD_Std = sd(CigsPerDay[Eversmoker == 1], na.rm = TRUE),
      PY_Avg = mean(Pack_Years[Eversmoker == 1], na.rm = TRUE),
      PY_Std = sd(Pack_Years[Eversmoker == 1], na.rm = TRUE),
      SY_Avg = mean(Smoking_Years[Eversmoker == 1], na.rm = TRUE),
      SY_Std = sd(Smoking_Years[Eversmoker == 1], na.rm = TRUE)
    ) %>%
    arrange(Age, Sex) %>%  # Sort by Age and Sex
    ungroup()
  
  # Return the summary dataframe ---------------------------------------------
  return(summary_df)
}
```


#dat2age
This function is designed specifically for the SCCS dataset. It transforms and prepares the data by:
Renaming key columns for consistency (Enrollment_Age to Age, etc.).
Recoding the Sex column (M → 0, F → 1).
Replacing missing values (NA) in Eversmoker and Smokenow with 0.
Calculating a new Smoking_Years column based on smoking history. Once the transformations are complete, the function calls smpl2age to generate a summarized dataframe showing smoking-related metrics by Age and Sex.

```{r}
dat2age <- function(data) {
  # Input Validation ---------------------------------------------------------
  if (!is.data.frame(data)) {
    stop("Input 'data' must be a dataframe.")
  }

  required_columns <- c("Enrollment_Age", "Sex", "RaceAnalysis", 
                        "EverSmoker", "SmokeNow", 
                        "PackYears", "AgeQuitSmoking", "AgeStartSmoking")
  missing_columns <- setdiff(required_columns, colnames(data))
  
  if (length(missing_columns) > 0) {
    stop(paste("Missing required columns:", paste(missing_columns, collapse = ", ")))
  }

  # Step 1: Replace NA in Eversmoker and Smokenow ----------------------------
  data <- data %>%
    mutate(
      EverSmoker = ifelse(is.na(EverSmoker), 0, EverSmoker),
      SmokeNow = ifelse(is.na(SmokeNow), 0, SmokeNow)
    )

  # Step 2: Rename columns ---------------------------------------------------
  data <- data %>%
    rename(
      Age = Enrollment_Age,
      Race = RaceAnalysis,
      Eversmoker = EverSmoker,
      Smokenow = SmokeNow,
      Pack_Years = PackYears
    )
  
  # Step 3: Recode Sex column ------------------------------------------------
  data <- data %>%
    mutate(Sex = case_when(
      Sex == "M" ~ 0,
      Sex == "F" ~ 1,
      TRUE ~ NA_real_
    ))

  # Step 4: Calculate Smoking_Years column -----------------------------------
  data <- data %>%
    mutate(Smoking_Years = case_when(
      Eversmoker == 0 ~ 0,
      Eversmoker == 1 & Smokenow == 0 ~ AgeQuitSmoking - AgeStartSmoking,
      Eversmoker == 1 & Smokenow == 1 ~ Age - AgeStartSmoking,
      TRUE ~ NA_real_
    ))

  # Step 5: Call smpl2age ----------------------------------------------------
  summary_df <- smpl2age(data)

  # Return the summary dataframe ---------------------------------------------
  return(summary_df)
}
```


#aggregate_density
This function aggregates density estimates from multiple samples for a given variable, such as CigsPerDay, Pack_Years, or Smoking_Years. It calculates the mean density and confidence interval (CI) bounds for each bin (x value), summarizing distributional trends across samples. This approach avoids storing large raw datasets while retaining essential information about variability and central tendencies in the sampled distributions.

```{r}
aggregate_density <- function(dist_list, var_name, CI = 0.95) {
  # Combine density data for all repetitions into a single dataframe
  all_densities <- bind_rows(
    lapply(dist_list, function(d) data.frame(x = d[[var_name]]$x, y = d[[var_name]]$y)),
    .id = "Rep"
  )
  
  # Aggregate densities by x-values
  aggregated <- all_densities %>%
    group_by(x) %>%
    summarise(
      Mean = mean(y),
      Lower = quantile(y, probs = (1 - CI) / 2),
      Upper = quantile(y, probs = 1 - (1 - CI) / 2),
      .groups = "drop"  # Drop grouping to avoid warnings
    )
  
  return(aggregated)
}
```


#rep_smpl_dist
This function performs repeated sampling from a dataset and tracks key variables (CigsPerDay, Pack_Years, Smoking_Years) and age-based smoking metrics (EverSmoker_Percentage, SmokeNow_Percentage). For each repetition, it calculates density distributions for ever-smokers and aggregates results into two dataframes: one with mean distributions and CI bounds, and another summarizing age-sex-specific smoking statistics with CI bounds. This function provides a robust framework for analyzing variability across samples.

```{r}
# Helper Function: Process Distributions for Eversmokers
process_distributions <- function(smpl) {
  smpl_ever <- smpl[smpl$Eversmoker == 1, ]
  
  list(
    CigsPerDay = density(smpl_ever$CigsPerDay, from = 0, to = 100, n = 512),
    Pack_Years = density(smpl_ever$Pack_Years, from = 0, to = 100, n = 512),
    Pack_Years_Minusone = density(smpl_ever$Pack_Years_Minusone, from = 0, to = 100, n = 512),
    Smoking_Years = density(smpl_ever$Smoking_Years, from = 0, to = 100, n = 512),
    Smoking_Years_Minusone = density(smpl_ever$Smoking_Years_Minusone, from = 0, to = 100, n = 512)
  )
}

# Main Function: rep_smpl_dist
rep_smpl_dist <- function(data, BYr, rep, CI = 0.95) {
  # Step 1: Initialize storage for distributions and age breakdown
  dist_list <- list(full = list(), female = list(), male = list())
  age_breakdown_list <- list()
  
  # Step 2: Main loop over repetitions
  for (seed in 1:rep) {
    # Sample once for the full population
    smpl_full <- sample_SHG(data, BYr, seed)
    
    # Filter subsets for female and male populations
    smpl_female <- smpl_full %>% filter(Sex == 1)
    smpl_male <- smpl_full %>% filter(Sex == 0)
    
    # Collect distributions for all three subpopulations
    dist_list$full[[seed]] <- process_distributions(smpl_full)
    dist_list$female[[seed]] <- process_distributions(smpl_female)
    dist_list$male[[seed]] <- process_distributions(smpl_male)
    
    # Age breakdown (only for full population)
    age_breakdown_list[[seed]] <- smpl2age(smpl_full)
  }
  
  # Step 3: Aggregate distributions across repetitions
  dist_aggregated <- list(
    full = list(
      CigsPerDay = aggregate_density(dist_list$full, "CigsPerDay", CI),
      Pack_Years = aggregate_density(dist_list$full, "Pack_Years", CI),
      Pack_Years_Minusone = aggregate_density(dist_list$full, "Pack_Years_Minusone", CI),
      Smoking_Years = aggregate_density(dist_list$full, "Smoking_Years", CI),
      Smoking_Years_Minusone = aggregate_density(dist_list$full, "Smoking_Years_Minusone", CI)
    ),
    female = list(
      CigsPerDay = aggregate_density(dist_list$female, "CigsPerDay", CI),
      Pack_Years = aggregate_density(dist_list$female, "Pack_Years", CI),
      Pack_Years_Minusone = aggregate_density(dist_list$female, "Pack_Years_Minusone", CI),
      Smoking_Years = aggregate_density(dist_list$female, "Smoking_Years", CI),
      Smoking_Years_Minusone = aggregate_density(dist_list$female, "Smoking_Years_Minusone", CI)
    ),
    male = list(
      CigsPerDay = aggregate_density(dist_list$male, "CigsPerDay", CI),
      Pack_Years = aggregate_density(dist_list$male, "Pack_Years", CI),
      Pack_Years_Minusone = aggregate_density(dist_list$male, "Pack_Years_Minusone", CI),
      Smoking_Years = aggregate_density(dist_list$male, "Smoking_Years", CI),
      Smoking_Years_Minusone = aggregate_density(dist_list$male, "Smoking_Years_Minusone", CI)
    )
  )
  
  # Step 4: Aggregate age breakdowns
  age_results <- bind_rows(age_breakdown_list, .id = "Rep") %>%
    group_by(Age, Sex) %>%
    summarise(
      EverSmoker_Mean = mean(EverSmoker_Percentage),
      EverSmoker_Lower = quantile(EverSmoker_Percentage, probs = (1 - CI) / 2),
      EverSmoker_Upper = quantile(EverSmoker_Percentage, probs = 1 - (1 - CI) / 2),
      SmokeNow_Mean = mean(SmokeNow_Percentage),
      SmokeNow_Lower = quantile(SmokeNow_Percentage, probs = (1 - CI) / 2),
      SmokeNow_Upper = quantile(SmokeNow_Percentage, probs = 1 - (1 - CI) / 2),
      .groups = "drop"
    )
  
  # Return aggregated results
  return(list(Distributions = dist_aggregated, AgeBreakdown = age_results))
}
```


#extr_prvlnc_smpld_rslts
The extr_prvlnc_smpld_rslts function extracts and formats the age-based prevalence and smoking metrics from the output of the rep_smpl_dist function. It focuses on the AgeBreakdown component of the input, ensuring the necessary columns (e.g., EverSmoker_Mean, SmokeNow_Mean, and their confidence intervals) are present and correctly formatted. The function directly selects these columns, reorders them for clarity, and arranges the data by age and sex. The resulting dataframe includes columns for age, sex, and smoking-related metrics with their mean and confidence interval bounds, making it ready for further analysis or visualization.

```{r}
extr_prvlnc_smpld_rslts <- function(smpld_SHG_rslts, CI = 0.95) {
  # Input Validation ---------------------------------------------------------
  if (!is.list(smpld_SHG_rslts) || !"AgeBreakdown" %in% names(smpld_SHG_rslts)) {
    stop("Input 'smpld_SHG_rslts' must be a list containing an 'AgeBreakdown' component.")
  }

  # Extract AgeBreakdown -----------------------------------------------------
  age_breakdown <- smpld_SHG_rslts$AgeBreakdown

  # Check for necessary columns ----------------------------------------------
  required_cols <- c("Age", "Sex", "EverSmoker_Mean", "SmokeNow_Mean", 
                     "EverSmoker_Lower", "EverSmoker_Upper", 
                     "SmokeNow_Lower", "SmokeNow_Upper")
  missing_cols <- setdiff(required_cols, colnames(age_breakdown))
  if (length(missing_cols) > 0) {
    stop(paste("Missing columns in 'AgeBreakdown':", paste(missing_cols, collapse = ", ")))
  }

  # Assign confidence interval columns directly ------------------------------
  result <- age_breakdown %>%
    select(
      Age, Sex,
      EverSmoker_Mean, 
      EverSmoker_Lower, EverSmoker_Upper,
      SmokeNow_Mean, 
      SmokeNow_Lower, SmokeNow_Upper
    ) %>%
    arrange(Age, Sex)  # Optional: ensure ordering by Age and Sex

  # Return the formatted dataframe -------------------------------------------
  return(result)
}
```


#extr_dist_smpld_rslts
The extr_dist_smpld_rslts function focuses on extracting and consolidating the distributions of smoking-related variables (CigsPerDay, Pack_Years, and Smoking_Years) from the Distributions component of the rep_smpl_dist output. It ensures that the input structure includes required columns (x, Mean, Lower, Upper) for each variable. The function combines these distributions into a single dataframe and reshapes it into a wide format, where each variable’s distribution is represented with its mean and confidence interval bounds. This structured output enables easy comparisons of variable distributions and is ready for detailed statistical analysis or visualization.

```{r}
extr_dist_smpld_rslts <- function(smpld_SHG_rslts, population = "full", CI = 0.95) {
  # Validate the population input
  if (!population %in% c("full", "female", "male")) {
    stop("Invalid population input. Choose from 'full', 'female', or 'male'.")
  }
  
  # Access the specific subpopulation distributions
  dist <- smpld_SHG_rslts$Distributions[[population]]
  
  # Combine distributions into a single dataframe in the desired format
  aggregated_dist <- tibble(
    x = dist$CigsPerDay$x,  # x-values are the same across all distributions
    CigsPerDay_Mean = dist$CigsPerDay$Mean,
    CigsPerDay_Lower = dist$CigsPerDay$Lower,
    CigsPerDay_Upper = dist$CigsPerDay$Upper,
    Pack_Years_Mean = dist$Pack_Years$Mean,
    Pack_Years_Lower = dist$Pack_Years$Lower,
    Pack_Years_Upper = dist$Pack_Years$Upper,
    Pack_Years_M1_Mean = dist$Pack_Years_Minusone$Mean,
    Pack_Years_M1_Lower = dist$Pack_Years_Minusone$Lower,
    Pack_Years_M1_Upper = dist$Pack_Years_Minusone$Upper,
    Smoking_Years_Mean = dist$Smoking_Years$Mean,
    Smoking_Years_Lower = dist$Smoking_Years$Lower,
    Smoking_Years_Upper = dist$Smoking_Years$Upper,
    Smoking_Years_M1_Mean = dist$Smoking_Years_Minusone$Mean,
    Smoking_Years_M1_Lower = dist$Smoking_Years_Minusone$Lower,
    Smoking_Years_M1_Upper = dist$Smoking_Years_Minusone$Upper
  )
  
  # Return the aggregated dataframe
  return(aggregated_dist)
}
```


#make_prvlnc_plot
Creates a prevalence plot comparing SHG simulations and data by age, sex, and smoking status. Includes confidence intervals, customizable legends, and titles. Inputs: SHG and data prevalence, target ("EverSmoker" or "SmokeNow"), CI level, and display options for legends and subtitles. Outputs a ggplot2 object.

```{r}
# Helper function for prevalence plots
make_prevalence_plot <- function(SHG_prvlnc, dat_prvlnc, target, CI = 95, legend = "On", subtitle = NULL) {
  # Validate inputs
  if (!target %in% c("EverSmoker", "SmokeNow")) {
    stop("Invalid target. Choose either 'EverSmoker' or 'SmokeNow'.")
  }
  required_cols <- c("Age", "Sex", paste0(target, "_Mean"), paste0(target, "_Lower"), paste0(target, "_Upper"))
  if (!all(required_cols %in% colnames(SHG_prvlnc))) {
    stop(paste("SHG_prvlnc must contain the necessary columns for", target, "analysis."))
  }
  if (!all(c("Age", "Sex", paste0(target, "_Percentage")) %in% colnames(dat_prvlnc))) {
    stop(paste("dat_prvlnc must contain the necessary columns for", target, "analysis."))
  }

  # Split data by gender
  SHG_male <- SHG_prvlnc[SHG_prvlnc$Sex == 0, ]
  SHG_female <- SHG_prvlnc[SHG_prvlnc$Sex == 1, ]
  dat_male <- dat_prvlnc[dat_prvlnc$Sex == 0, ]
  dat_female <- dat_prvlnc[dat_prvlnc$Sex == 1, ]

  # Define labels for legend
  CI_label_m <- paste0("Males (", CI, "% CI)")
  CI_label_f <- paste0("Females (", CI, "% CI)")
  y_label <- paste(target, "Percentage")

  # Set up plot title
  plottitle <- if (!is.null(subtitle)) subtitle else NULL

  # Create the plot
  plot <- ggplot() +
    geom_ribbon(
      data = SHG_male,
      aes(x = Age, ymin = .data[[paste0(target, "_Lower")]], ymax = .data[[paste0(target, "_Upper")]], fill = CI_label_m),
      alpha = 0.4
    ) +
    geom_ribbon(
      data = SHG_female,
      aes(x = Age, ymin = .data[[paste0(target, "_Lower")]], ymax = .data[[paste0(target, "_Upper")]], fill = CI_label_f),
      alpha = 0.4
    ) +
    geom_line(
      data = SHG_male,
      aes(x = Age, y = .data[[paste0(target, "_Mean")]], color = "Males (SHG)"),
      size = 1
    ) +
    geom_line(
      data = SHG_female,
      aes(x = Age, y = .data[[paste0(target, "_Mean")]], color = "Females (SHG)"),
      size = 1
    ) +
    geom_point(
      data = dat_male,
      aes(x = Age, y = .data[[paste0(target, "_Percentage")]], color = "Males (Data)"),
      size = 2
    ) +
    geom_point(
      data = dat_female,
      aes(x = Age, y = .data[[paste0(target, "_Percentage")]], color = "Females (Data)"),
      size = 2
    ) +
    labs(
      title = plottitle,
      x = "Age",
      y = y_label,
      fill = "Legend",
      color = "Legend"
    ) +
    scale_color_manual(
      values = c(
        "Females (Data)" = "darkred",
        "Females (SHG)" = "red",
        "Males (Data)" = "darkblue",
        "Males (SHG)" = "blue"
      ),
      breaks = c("Females (Data)", "Females (SHG)", "Males (Data)", "Males (SHG)")  # Controls order in the legend
    ) +
    scale_fill_manual(
      values = setNames(c("pink", "lightblue"), c(CI_label_f, CI_label_m)),  # Ensure mapping matches dynamic labels
      breaks = c(CI_label_f, CI_label_m)  # Specify order in the legend
    ) +
    coord_cartesian(ylim = c(0, 100)) +
    theme_minimal() +
    theme(
      legend.position = if (legend == "On") "bottom" else "none",
      legend.box = "horizontal",
      legend.title = element_blank(),
      plot.title = element_text(hjust = 0.5, size = 13)
    ) +
    guides(
      color = guide_legend(nrow = 2, byrow = TRUE),  # Two-row legend for lines
      fill = guide_legend(nrow = 2, byrow = TRUE)    # Two-row legend for ribbons
    )

  return(plot)
}
```


#prvlnc_plot
Generates two side-by-side plots comparing SHG-simulated and observed smoking prevalence by age and sex for Ever Smokers and Current Smokers. Each plot shows confidence intervals, mean prevalence lines, and data points for males and females. Ensures consistent y-axis limits, distinct color coding, and a single shared legend. Validates input data for required columns to ensure reliable execution.

```{r}
prvlnc_plot <- function(SHG_sim_prvlnc, dat_prvlnc, 
                        legend = "On", subtitle = NULL, population = NULL, CI = 95) {
  # Validate inputs for required columns
  required_cols <- c("Age", "Sex", "EverSmoker_Mean", "EverSmoker_Lower", "EverSmoker_Upper",
                     "SmokeNow_Mean", "SmokeNow_Lower", "SmokeNow_Upper")
  if (!all(required_cols %in% colnames(SHG_sim_prvlnc))) {
    stop("SHG_sim_prvlnc must contain the necessary columns for prevalence analysis.")
  }
  if (!all(c("Age", "Sex", "EverSmoker_Percentage", "SmokeNow_Percentage") %in% colnames(dat_prvlnc))) {
    stop("dat_prvlnc must contain the necessary columns for prevalence analysis.")
  }
  
  # Set up subtitles for the individual plots
  sub1 <- if (!is.null(subtitle)) paste0("Ever Smokers of ", population) else population
  sub2 <- if (!is.null(subtitle)) paste0("Current Smokers of ", population) else population
  
  # Generate the Ever Smokers plot
  plot_ever <- make_prevalence_plot(
    SHG_prvlnc = SHG_sim_prvlnc,
    dat_prvlnc = dat_prvlnc,
    target = "EverSmoker",
    CI = CI,
    legend = legend,
    subtitle = sub1
  )

  # Generate the Current Smokers plot
  plot_now <- make_prevalence_plot(
    SHG_prvlnc = SHG_sim_prvlnc,
    dat_prvlnc = dat_prvlnc,
    target = "SmokeNow",
    CI = CI,
    legend = "Off",  # Suppress individual legends for the second plot
    subtitle = sub2
  )

  # Combine everything with one shared legend
  final_plot <- (plot_ever | plot_now) +
    plot_layout(guides = "collect") +  # Collect all legends into one
    plot_annotation(
      theme = theme(
        legend.position = if (legend == "On") "bottom" else "none",
        legend.box = "horizontal",
        legend.title = element_blank()
      )
    )
  
  # Return the final combined plot
  return(final_plot)
}
```


#dist_plot
The dist_plot function visualizes and compares the distributions of SHG simulation results and observed data for a specified target variable (CigsPerDay, PackYears, or SmokingYears). It plots the SHG mean line with a confidence interval ribbon and overlays the density of the observed data for EverSmokers. The function outputs a ggplot object with a legend distinguishing SHG CI, SHG mean, and observed data, making it easy to assess alignment between simulations and observations.

```{r}
dist_plot <- function(dist_smpled, data, target, 
                      title = "Comparison of Distributions: SHG vs. Data",
                      legend = "On",
                      ytitle = "On",
                      ymax = 0,
                      CI = 95) {
  # Validate input for target
  if (!target %in% c("CigsPerDay", "PackYears", "SmokingYears")) {
    stop("Invalid target. Choose from 'CigsPerDay', 'PackYears', or 'SmokingYears'.")
  }
  
  # Determine columns based on target
  target_cols <- switch(target,
                        CigsPerDay = c("CigsPerDay_Mean", "CigsPerDay_Lower", "CigsPerDay_Upper"),
                        PackYears = c("Pack_Years_Mean", "Pack_Years_Lower", "Pack_Years_Upper"),
                        SmokingYears = c("Smoking_Years_Mean", "Smoking_Years_Lower", "Smoking_Years_Upper"))
  
  # Map x-axis label
  x_label <- switch(target,
                    CigsPerDay = "Cigarettes Per Day",
                    PackYears = "Pack-Years",
                    SmokingYears = "Smoking Duration (Years)")
  
  # Extract relevant data
  target_data <- data[[target]][data$EverSmoker == 1]
  data_density <- density(target_data[!is.na(target_data)], from = 0, to = 100, n = 512)
  data_dist <- data.frame(
    x = data_density$x,
    y = data_density$y
  )
  sampled_mean <- dist_smpled[[target_cols[1]]]
  sampled_lower <- dist_smpled[[target_cols[2]]]
  sampled_upper <- dist_smpled[[target_cols[3]]]
  
  # Determine ymax if not specified
  if (ymax == 0) {
    ymax <- max(data_density$y, sampled_upper, na.rm = TRUE)
  }
  
  # Prepare CI label
  CI_label <- paste0(CI, "% CI")
  
  # Create a data frame for plotting
  plot_data <- data.frame(
    x = data_dist$x,
    SCCS = data_dist$y,
    SHG = sampled_mean,
    SHG_Lower = sampled_lower,
    SHG_Upper = sampled_upper
  )
  
  # Plot using ggplot2
  Plot_dist <- ggplot(plot_data) +
    # Confidence ribbon (SHG CI)
    geom_ribbon(
      aes(x = x, ymin = SHG_Lower, ymax = SHG_Upper, fill = CI_label),
      alpha = 0.8
    ) +
    # Mean line (SHG)
    geom_line(
      aes(x = x, y = SHG, color = "SHG"),
      size = 1
    ) +
    # Observed SCCS data line
    geom_line(
      aes(x = x, y = SCCS, color = "Data"),
      size = 1
    ) +
    # Axis limits, labels, and theme
    coord_cartesian(xlim = c(0, 80), ylim = c(0, ymax)) +
    labs(
      title = title,
      x = x_label,
      y = if (ytitle == "On") "Density" else NULL,
      fill = "Legend",
      color = "Legend"
    ) +
    # Custom legend with controlled order
    guides(
      color = if (legend == "On") guide_legend(order = 1, override.aes = list(size = 1)) else "none",
      fill = if (legend == "On") guide_legend(order = 2, override.aes = list(alpha = 0.8)) else "none"
    ) +
    scale_color_manual(
      values = c("Data" = "orange", "SHG" = "blue"),
      breaks = c("Data", "SHG")
    ) +
    scale_fill_manual(
      values = setNames("lightblue", CI_label),
      breaks = c(CI_label)
    ) +
    theme_minimal() +
    theme(
      legend.box.background = if (legend == "On") element_rect(color = "black", size = 0.5) else element_blank(),
      legend.box.margin = if (legend == "On") margin(5, 5, 5, 5) else margin(0, 0, 0, 0),
      legend.key.size = if (legend == "On") unit(1, "lines") else unit(0, "lines"),
      legend.title = element_blank()
    )

  return(Plot_dist)
}
```


#stacked_dist_plots
The stacked_dist_plots function generates and combines distribution comparison plots for CigsPerDay, PackYears, and SmokingYears. It calls the dist_plot function three times, once for each target, and stacks the resulting plots vertically using the patchwork package. The combined plot includes individual titles for each variable as well as an overarching title, providing a clear visualization of the SHG simulations compared to the observed data across key smoking-related variables.

```{r}
# Function to create and stack distribution plots
stacked_dist_plots <- function(dist_smpled, data, 
                               title = "SHG vs. Data: Distributions of Smoking Variables",
                               legend = "On",
                               subtitle = "On",
                               ytitle = "On",
                               ymax = c(0, 0, 0),
                               CI = 95,
                               legend_position = c("last", "all", "none")) {
  
  legend_position <- match.arg(legend_position)

  # Determine subplot titles based on subtitle parameter
  cigs_title <- if (subtitle == "On") "Cigarettes Per Day: SHG vs. Data" else NULL
  packyears_title <- if (subtitle == "On") "Pack Years: SHG vs. Data" else NULL
  smokingyears_title <- if (subtitle == "On") "Smoking Years: SHG vs. Data" else NULL

  # Determine legend settings per plot
  legend_flags <- switch(
    legend_position,
    "last" = c("Off", "Off", legend),
    "all" = rep(legend, 3),
    "none" = rep("Off", 3)
  )

  # Generate individual plots for each target
  plot_cigs <- dist_plot(
    dist_smpled = dist_smpled,
    data = data,
    target = "CigsPerDay",
    title = cigs_title,
    legend = legend_flags[1],
    ytitle = ytitle,
    ymax = ymax[1],
    CI = CI
  )
  
  plot_packyears <- dist_plot(
    dist_smpled = dist_smpled,
    data = data,
    target = "PackYears",
    title = packyears_title,
    legend = legend_flags[2],
    ytitle = ytitle,
    ymax = ymax[2],
    CI = CI
  )
  
  plot_smokingyears <- dist_plot(
    dist_smpled = dist_smpled,
    data = data,
    target = "SmokingYears",
    title = smokingyears_title,
    legend = legend_flags[3],
    ytitle = ytitle,
    ymax = ymax[3],
    CI = CI
  )
  
  # Stack the plots using patchwork
  combined_plot <- plot_cigs / plot_packyears / plot_smokingyears +
    plot_layout(heights = c(1, 1, 1), guides = "collect") +
    plot_annotation(title)

  # Adjust the legend position to the bottom
  combined_plot <- combined_plot & 
    theme(legend.position = "bottom")

  return(combined_plot)
}
```


#calculate_ymax
This function computes the maximum y-axis value (ymax) for a smoking variable by comparing the highest density values from observed data (data_1, data_2) and simulated distributions (dist_smpled_1, dist_smpled_2).
Inputs: Simulated distributions (dist_smpled_1, dist_smpled_2), Observed data (data_1, data_2), Target variable ("CigsPerDay", "PackYears", or "SmokingYears")
Output: The maximum y-axis value (ymax).

```{r}
calculate_ymax <- function(dist_smpled_1, dist_smpled_2, data_1, data_2, target) {
  # Determine column names for dist_smpled based on the target
  target_upper <- paste0(target, ifelse(target %in% c("CigsPerDay"), "_Upper", "_Upper"))
  
  # Handle naming inconsistencies
  if (!(target_upper %in% colnames(dist_smpled_1))) {
    target_upper <- paste0(gsub("Years", "_Years", target), "_Upper")
  }
  
  max_values <- c(
    max(dist_smpled_1[[target_upper]], na.rm = TRUE),
    max(dist_smpled_2[[target_upper]], na.rm = TRUE),
    max(density(data_1[[target]][data_1$EverSmoker == 1 & !is.na(data_1[[target]])], 
                from = 0, to = 100, n = 512)$y, na.rm = TRUE),
    max(density(data_2[[target]][data_2$EverSmoker == 1 & !is.na(data_2[[target]])], 
                from = 0, to = 100, n = 512)$y, na.rm = TRUE)
  )
  
  if (all(is.infinite(max_values))) return(0)  # Fallback to 0 if no valid data
  return(max(max_values, na.rm = TRUE))
}
```


#stacked_side_plots
This function generates side-by-side stacked plots comparing smoking variable distributions for two populations. Each column includes subplots for "CigsPerDay", "PackYears", and "SmokingYears", with simulation and data results combined.
Inputs: Simulated distributions (dist_smpled_1, dist_smpled_2), Observed data (data_1, data_2), Optional column titles (column_titles) and overall title (maintitle).
Output: A combined ggplot object showing side-by-side stacked plots.

```{r}
stacked_side_plots <- function(dist_smpled_1, dist_smpled_2, data_1, data_2, 
                               maintitle = "Smoking History Distributions: Data vs. SHG", 
                               column_titles = c("Population 1", "Population 2"),
                               CI = 95) {
  # Calculate ymax for each variable
  ymax_cigs <- calculate_ymax(dist_smpled_1, dist_smpled_2, data_1, data_2, "CigsPerDay")
  ymax_packyears <- calculate_ymax(dist_smpled_1, dist_smpled_2, data_1, data_2, "PackYears")
  ymax_smokingyears <- calculate_ymax(dist_smpled_1, dist_smpled_2, data_1, data_2, "SmokingYears")
  ymax <- c(ymax_cigs, ymax_packyears, ymax_smokingyears)

  # Step 1: Temporarily generate plot_1 with a visible legend to extract it
  plot_1_with_legend <- stacked_dist_plots(
    dist_smpled = dist_smpled_1, 
    data = data_1, 
    title = NULL,
    legend = "On", 
    subtitle = "Off",
    ymax = ymax,
    CI = CI,
    legend_position = "last"  # keep legend only in the final panel
  )

  # Step 2: Extract the legend using cowplot
  legend_shared <- get_legend(plot_1_with_legend)

  # Step 3: Regenerate plot_1 and plot_2 with legends off
  plot_1 <- stacked_dist_plots(
    dist_smpled = dist_smpled_1, 
    data = data_1, 
    title = NULL,
    legend = "Off", 
    subtitle = "Off",
    ymax = ymax,
    CI = CI,
    legend_position = "none"
  )

  plot_2 <- stacked_dist_plots(
    dist_smpled = dist_smpled_2, 
    data = data_2, 
    title = NULL,
    legend = "Off", 
    subtitle = "Off",
    ytitle = "Off",
    ymax = ymax,
    CI = CI,
    legend_position = "none"
  )

  # Create the column titles
  title_main <- ggdraw() + 
    draw_label(maintitle, fontface = "bold", size = 13, hjust = 0.5)

  title_row <- plot_grid(
    ggdraw() + draw_label(column_titles[1], hjust = 0.5, size = 11),
    ggdraw() + draw_label(column_titles[2], hjust = 0.5, size = 11),
    ncol = 2
  )

  # Combine plot columns side-by-side
  plot_body <- plot_grid(plot_1, plot_2, ncol = 2, rel_widths = c(1, 1))

  # Final layout: title -> column titles -> plots -> legend
  final_plot <- plot_grid(
    title_main,
    title_row,
    plot_body,
    legend_shared,
    ncol = 1,
    rel_heights = c(0.04, 0.06, 1, 0.12)
  )

  return(final_plot)
}
```


# dist_plot_2
This function compares the observed SCCS density for one smoking-history variable against the corresponding SHG mean density and confidence interval. It returns a single ggplot object with consistent axis handling, legend controls, and optional shared y-axis scaling for use in stacked figures.

```{r}
dist_plot_2 <- function(dist_smpled, data, target, 
                      title = "Comparison of Distributions: SHG vs. Data",
                      legend = "On",
                      ytitle = "On",
                      ymax = 0,
                      CI = 95) {
  
  # Validate input
  if (!target %in% c("CigsPerDay", "PackYears", "SmokingYears")) {
    stop("Invalid target. Choose from 'CigsPerDay', 'PackYears', or 'SmokingYears'.")
  }
  
  # Map to SHG column names
  target_cols <- switch(target,
                        CigsPerDay = c("CigsPerDay_Mean", "CigsPerDay_Lower", "CigsPerDay_Upper"),
                        PackYears = c("Pack_Years_Mean", "Pack_Years_Lower", "Pack_Years_Upper"),
                        SmokingYears = c("Smoking_Years_Mean", "Smoking_Years_Lower", "Smoking_Years_Upper"))
  
  # x-axis label
  x_label <- switch(target,
                    CigsPerDay = "Cigarettes Per Day",
                    PackYears = "Pack-Years",
                    SmokingYears = "Smoking Duration (Years)")
  
  # Prepare observed density
  target_data <- data[[target]][data$EverSmoker == 1]
  data_density <- density(target_data[!is.na(target_data)],
                          from = 0, to = 100, n = 512)  # ensure consistent x-range
  data_dist <- data.frame(x = data_density$x, y = data_density$y)
  
  # Extract SHG mean and CI
  sampled_mean  <- dist_smpled[[target_cols[1]]]
  sampled_lower <- dist_smpled[[target_cols[2]]]
  sampled_upper <- dist_smpled[[target_cols[3]]]
  
  # Prepare plotting data frame
  plot_data <- data.frame(
    x = data_density$x,  # ensure shared x-grid
    SCCS      = data_dist$y,
    SHG       = sampled_mean,
    SHG_Lower = sampled_lower,
    SHG_Upper = sampled_upper
  )
  
  # Compute shared ymax if not provided
  if (ymax == 0) {
    ymax <- max(plot_data$SCCS, plot_data$SHG_Upper, na.rm = TRUE)
  }
  
  CI_label <- paste0(CI, "% CI")
  
  # Plot
  p <- ggplot(plot_data, aes(x = x)) +
    # Confidence ribbon
    geom_ribbon(aes(ymin = SHG_Lower, ymax = SHG_Upper, fill = CI_label), alpha = 0.8) +
    # SHG mean line
    geom_line(aes(y = SHG, color = "SHG"), linewidth = 1.3) +
    # Observed data line
    geom_line(aes(y = SCCS, color = "SCCS"), linewidth = 1.3) +
    
    # Axis scaling and formatting
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(
      expand = expansion(mult = c(0, 0.02)),
      limits = c(0, ymax),
      labels = scales::label_number(accuracy = 0.01)
    ) +
    
    coord_cartesian(xlim = c(0, 80), ylim = c(0, ymax)) +
    
    labs(
      title = title,
      x = x_label,
      y = if (ytitle == "On") "Density" else NULL,
      fill = "Legend",
      color = "Legend"
    ) +
    
    # Custom legend and color styling
    scale_color_manual(
      values = c("SCCS" = "orange", "SHG" = "blue"),
      breaks = c("SCCS", "SHG"),
      labels = c("SCCS (observed)", "SHG (simulated)")
    ) +
    scale_fill_manual(
      values = setNames("lightblue", CI_label),
      breaks = c(CI_label)
    ) +
    
    guides(
      color = if (legend == "On") guide_legend(order = 1) else "none",
      fill  = if (legend == "On") guide_legend(order = 2) else "none"
    ) +
    
    theme_minimal(base_size = 16) +
    theme(
      plot.title      = element_text(size = 18, face = "bold", hjust = 0.5),
      axis.title.x    = element_text(size = 16),
      axis.title.y    = element_text(size = 16),
      axis.text       = element_text(size = 14),
      legend.text     = element_text(size = 14),
      legend.title    = element_blank(),
      legend.position = if (legend == "On") "bottom" else "none",
      axis.line       = element_line(color = "black"),
      panel.border    = element_blank(),
      plot.margin     = margin(5, 5, 5, 5)
    )
  
  return(p)
}
```


#stacked_dist_plots_2
This function builds the three single-variable distribution comparisons from dist_plot_2 and returns them as a named list. It allows the caller to control subtitles, legend placement, shared y-axis scaling, and whether the x-axis is shown on each component plot.

```{r}
stacked_dist_plots_2 <- function(dist_smpled, data, 
                             subtitle = "On",
                             ytitle = "On",
                             ymax = c(0, 0, 0),
                             CI = 95,
                             legend_flags = c("Off", "Off", "Off"),
                             show_x_axis = c(TRUE, TRUE, TRUE)) {
  # Axis suppressor
  strip_x <- function(p) {
    p + theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
    ) + xlab(NULL)
  }

  # Titles
  titles <- if (subtitle == "On") {
    c("Cigarettes Per Day: SHG vs. Data", "Pack Years: SHG vs. Data", "Smoking Years: SHG vs. Data")
  } else {
    rep(NULL, 3)
  }

  # Cigs
  plot_cigs <- dist_plot_2(dist_smpled, data, "CigsPerDay", titles[1], legend_flags[1], ytitle, ymax[1], CI)
  if (!show_x_axis[1]) plot_cigs <- strip_x(plot_cigs)

  # Pack Years
  plot_packyears <- dist_plot_2(dist_smpled, data, "PackYears", titles[2], legend_flags[2], ytitle, ymax[2], CI)
  if (!show_x_axis[2]) plot_packyears <- strip_x(plot_packyears)

  # Smoking Years
  plot_smokingyears <- dist_plot_2(dist_smpled, data, "SmokingYears", titles[3], legend_flags[3], ytitle, ymax[3], CI)
  if (!show_x_axis[3]) plot_smokingyears <- strip_x(plot_smokingyears)

  return(list(
    cigs = plot_cigs,
    packyears = plot_packyears,
    smokingyears = plot_smokingyears
  ))
}
```



#stacked_plots
This function arranges two groups of SCCS-vs-SHG distribution plots into a single multi-row figure with optional shared title and legend. It standardizes y-axis limits across the paired groups so that differences between rows can be compared on the same scale.

```{r}
stacked_plots <- function(
  dist_smpled_1, dist_smpled_2, data_1, data_2, 
  maintitle = "Smoking History Distributions: SHG vs. Data", 
  row_titles = c("Group 1", "Group 2"),
  CI = 95,
  include_title = TRUE,
  include_legend = TRUE
) {
  # Calculate ymax
  ymax <- c(
    calculate_ymax(dist_smpled_1, dist_smpled_2, data_1, data_2, "CigsPerDay"),
    calculate_ymax(dist_smpled_1, dist_smpled_2, data_1, data_2, "PackYears"),
    calculate_ymax(dist_smpled_1, dist_smpled_2, data_1, data_2, "SmokingYears")
  )

  # Extract legend if needed
  if (include_legend) {
    legend_plot <- dist_plot_2(
      dist_smpled = dist_smpled_1,
      data = data_1,
      target = "SmokingYears",
      title = NULL,
      legend = "On",
      ytitle = "Off",
      ymax = ymax[3],
      CI = CI
    ) + theme(
      legend.text = element_text(size = 20),
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.background = element_blank(),
      legend.box.background = element_blank(),
      panel.background = element_blank(),
      plot.background = element_blank()
    )
    legend_shared <- get_legend(legend_plot)
  }

  # Top row (no x-axis)
  plots_pop1 <- stacked_dist_plots_2(
    dist_smpled_1, data_1,
    subtitle = "Off", ytitle = "Off", ymax = ymax, CI = CI,
    legend_flags = c("Off", "Off", "Off"),
    show_x_axis = c(FALSE, FALSE, FALSE)
  )

  # Bottom row (with x-axis)
  plots_pop2 <- stacked_dist_plots_2(
    dist_smpled_2, data_2,
    subtitle = "Off", ytitle = "Off", ymax = ymax, CI = CI,
    legend_flags = c("Off", "Off", "Off"),
    show_x_axis = if (include_legend) c(TRUE, TRUE, TRUE) else c(FALSE, FALSE, FALSE)
  )

  # Rows of plots
  row1 <- plot_grid(plots_pop1$cigs, plots_pop1$packyears, plots_pop1$smokingyears, ncol = 3)
  row2 <- plot_grid(plots_pop2$cigs, plots_pop2$packyears, plots_pop2$smokingyears, ncol = 3)

  # Row labels (rotated inward)
  row_label_1 <- ggdraw() + draw_label(row_titles[1], angle = 270, vjust = 0.5, hjust = 0.5, size = 20)
  row_label_2 <- ggdraw() + draw_label(row_titles[2], angle = 270, vjust = 0.5, hjust = 0.5, size = 20)

  # Add row labels to right
  row1_labeled <- plot_grid(row1, row_label_1, ncol = 2, rel_widths = c(1, 0.05), align = "v")
  row2_labeled <- plot_grid(row2, row_label_2, ncol = 2, rel_widths = c(1, 0.05), align = "v")

  # Stack rows
  plot_body <- plot_grid(
    row1_labeled,
    row2_labeled,
    ncol = 1,
    rel_heights = if (include_legend) c(1, 1.3) else c(1, 1),
    align = "v"
  )
  # Title and layout
  title_main <- ggdraw() + draw_label(maintitle, fontface = "bold", size = 22, hjust = 0.5)

  plot_list <- list()
  if (include_title) plot_list <- c(plot_list, list(title_main))
  plot_list <- c(plot_list, list(plot_body))
  if (include_legend) plot_list <- c(plot_list, list(legend_shared))

  rel_heights <- c()
  if (include_title) rel_heights <- c(rel_heights, 0.06)
  rel_heights <- c(rel_heights, 1)
  if (include_legend) rel_heights <- c(rel_heights, 0.12)

  final_plot <- plot_grid(plotlist = plot_list, ncol = 1, rel_heights = rel_heights)
  return(final_plot)
}
```


#count_eligible
The count_eligible function calculates the number of individuals meeting smoking screening criteria based on age and smoking history. It filters the dataset to include those within a specified age range who meet or exceed a given threshold for a smoking measure (e.g., Pack Years or Smoking Years). It considers both current smokers and former smokers who quit within the last 15 years (YSQ ≤ 15), returning the total count of eligible individuals.

```{r}
# Helper function for eligibility criteria
count_eligible <- function(df, measure, threshold, min_age, max_age = 80, 
                           age_col, smoker_col) {
  df %>%
    filter(.data[[age_col]] >= min_age, .data[[age_col]] <= max_age, 
           .data[[measure]] >= threshold, 
           (.data[[smoker_col]] == 1 | (.data[[smoker_col]] == 0 & YSQ <= 15))) %>%
    nrow()
}
``` 


#screening_eligible
The screening_eligible function calculates key smoking-related statistics for a given dataset, distinguishing between simulated SHG data and real-world SCCS data. It computes the total population, the number of ever-smokers, current smokers, and former smokers. Additionally, it determines the prevalence of individuals exceeding pack-year and smoking-year thresholds (PY≥20, PY≥30, SY≥20, SY≥30) and assesses eligibility for lung cancer screening based on 2021 and 2013 USPSTF guidelines, as well as alternative SY-based criteria. The function returns these values in a structured dataframe.

```{r}
screening_eligible <- function(df, dataset_type = "SHG") {
  # Validate dataset_type
  if (!dataset_type %in% c("SHG", "SCCS")) {
    stop("Invalid dataset_type. Use 'SHG' for simulated data or 'SCCS' for real data.")
  }

  # Define column mappings for SHG vs SCCS
  if (dataset_type == "SHG") {
    age_col <- "Age"
    smoker_col <- "Smokenow"
    ever_col <- "Eversmoker"
    cessation_col <- "Cessation_Age"
    py_col <- "Pack_Years"
    sy_col <- "Smoking_Years"
  } else {  # dataset_type == "SCCS"
    age_col <- "Enrollment_Age"
    smoker_col <- "SmokeNow"
    ever_col <- "EverSmoker"
    cessation_col <- "AgeQuitSmoking"
    py_col <- "PackYears"
    sy_col <- "SmokingYears"
  }

  # Compute YSQ (Years Since Quitting)
  df <- df %>%
    mutate(YSQ = case_when(
      .data[[smoker_col]] == 1 ~ 0,  # Current smokers have YSQ = 0
      .data[[smoker_col]] == 0 & .data[[cessation_col]] > 0 ~ .data[[age_col]] - .data[[cessation_col]],  # Ex-smokers
      .data[[smoker_col]] == 0 & .data[[cessation_col]] <= 0 ~ NA_real_  # Invalid cessation age
    ))

  # Count populations
  total_population <- nrow(df)
  ever_smokers <- df %>% filter(.data[[ever_col]] == 1)
  count_eversmokers <- nrow(ever_smokers)
  count_currentsmokers <- nrow(ever_smokers %>% filter(.data[[smoker_col]] == 1))
  count_formersmokers <- count_eversmokers - count_currentsmokers  # Since EverSmoker = Current + Former

  # Compute prevalence of PY≥20, PY≥30, SY≥20, SY≥30
  prevalence_counts <- c(
    PY20 = nrow(ever_smokers %>% filter(.data[[py_col]] >= 20)),
    PY30 = nrow(ever_smokers %>% filter(.data[[py_col]] >= 30)),
    SY20 = nrow(ever_smokers %>% filter(.data[[sy_col]] >= 20)),
    SY30 = nrow(ever_smokers %>% filter(.data[[sy_col]] >= 30))
  )

  # Compute eligibility counts using the external helper function
  eligibility_counts <- c(
    Eligible_2021 = count_eligible(df = ever_smokers, 
                                        measure = py_col, 
                                        threshold = 20, 
                                        min_age = 50,
                                        max_age = 80,
                                        age_col = age_col,
                                        smoker_col = smoker_col),
    Eligible_2013 = count_eligible(df = ever_smokers, 
                                        measure = py_col, 
                                        threshold = 30, 
                                        min_age = 55,
                                        max_age = 80,
                                        age_col = age_col,
                                        smoker_col = smoker_col),
    Eligible_SY20 = count_eligible(df = ever_smokers, 
                                   measure = sy_col, 
                                   threshold = 20, 
                                   min_age = 50,
                                   max_age = 80,
                                   age_col = age_col,
                                   smoker_col = smoker_col),
    Eligible_SY30 = count_eligible(df = ever_smokers, 
                                   measure = sy_col, 
                                   threshold = 30, 
                                   min_age = 50,
                                   max_age = 80,
                                   age_col = age_col,
                                   smoker_col = smoker_col)
  )

  # Return structured dataframe
  return(data.frame(
    Dataset = dataset_type,
    Total_Population = total_population,
    EverSmokers = count_eversmokers,
    CurrentSmokers = count_currentsmokers,
    FormerSmokers = count_formersmokers,
    PY20 = prevalence_counts["PY20"],
    PY30 = prevalence_counts["PY30"],
    SY20 = prevalence_counts["SY20"],
    SY30 = prevalence_counts["SY30"],
    Eligible_2021 = eligibility_counts["Eligible_2021"],
    Eligible_2013 = eligibility_counts["Eligible_2013"],
    Eligible_SY20 = eligibility_counts["Eligible_SY20"],
    Eligible_SY30 = eligibility_counts["Eligible_SY30"]
  ))
}
```


#aggregate_screening
The aggregate_screening function processes repeated screening eligibility assessments across multiple samples and summarizes key statistics. It calculates the mean, lower, and upper confidence bounds for total population, ever-smokers, current smokers, former smokers, and individuals exceeding specific pack-year (PY) and smoking-year (SY) thresholds. Additionally, it aggregates eligibility estimates based on 2021 and 2013 USPSTF criteria, as well as SY-based alternatives. The function returns a structured dataframe with the population name as the first column.

```{r}
aggregate_screening <- function(screening_data, subpop_name, CI = 0.95) {
  df <- bind_rows(screening_data)
  
  df_summary <- df %>%
    summarise(
      Total_Population = mean(Total_Population),
      
      EverSmokers_Mean = mean(EverSmokers),
      EverSmokers_Lower = quantile(EverSmokers, probs = (1 - CI) / 2),
      EverSmokers_Upper = quantile(EverSmokers, probs = 1 - (1 - CI) / 2),
      
      CurrentSmokers_Mean = mean(CurrentSmokers),
      CurrentSmokers_Lower = quantile(CurrentSmokers, probs = (1 - CI) / 2),
      CurrentSmokers_Upper = quantile(CurrentSmokers, probs = 1 - (1 - CI) / 2),
      
      FormerSmokers_Mean = mean(FormerSmokers),
      FormerSmokers_Lower = quantile(FormerSmokers, probs = (1 - CI) / 2),
      FormerSmokers_Upper = quantile(FormerSmokers, probs = 1 - (1 - CI) / 2),
      
      PY20_Mean = mean(PY20),
      PY20_Lower = quantile(PY20, probs = (1 - CI) / 2),
      PY20_Upper = quantile(PY20, probs = 1 - (1 - CI) / 2),

      PY30_Mean = mean(PY30),
      PY30_Lower = quantile(PY30, probs = (1 - CI) / 2),
      PY30_Upper = quantile(PY30, probs = 1 - (1 - CI) / 2),

      SY20_Mean = mean(SY20),
      SY20_Lower = quantile(SY20, probs = (1 - CI) / 2),
      SY20_Upper = quantile(SY20, probs = 1 - (1 - CI) / 2),

      SY30_Mean = mean(SY30),
      SY30_Lower = quantile(SY30, probs = (1 - CI) / 2),
      SY30_Upper = quantile(SY30, probs = 1 - (1 - CI) / 2),

      Eligible_2021_Mean = mean(Eligible_2021),
      Eligible_2021_Lower = quantile(Eligible_2021, probs = (1 - CI) / 2),
      Eligible_2021_Upper = quantile(Eligible_2021, probs = 1 - (1 - CI) / 2),

      Eligible_2013_Mean = mean(Eligible_2013),
      Eligible_2013_Lower = quantile(Eligible_2013, probs = (1 - CI) / 2),
      Eligible_2013_Upper = quantile(Eligible_2013, probs = 1 - (1 - CI) / 2),

      Eligible_SY20_Mean = mean(Eligible_SY20),
      Eligible_SY20_Lower = quantile(Eligible_SY20, probs = (1 - CI) / 2),
      Eligible_SY20_Upper = quantile(Eligible_SY20, probs = 1 - (1 - CI) / 2),

      Eligible_SY30_Mean = mean(Eligible_SY30),
      Eligible_SY30_Lower = quantile(Eligible_SY30, probs = (1 - CI) / 2),
      Eligible_SY30_Upper = quantile(Eligible_SY30, probs = 1 - (1 - CI) / 2)
    ) %>%
    mutate(Population = subpop_name) %>%
    select(Population, everything())  # Ensure Population is first column

  return(df_summary)
}
```


#rep_smpl_screening
The rep_smpl_screening function performs repeated sampling from a simulated population and evaluates screening eligibility across different subpopulations. It draws multiple random samples (full population) and extracts male and female subsets within each iteration. Screening eligibility metrics are calculated separately for each group and stored. After all repetitions, the results are aggregated using aggregate_screening, summarizing mean estimates and confidence intervals for key measures. The function returns a structured dataframe with screening eligibility statistics for the full, male, and female populations.

```{r}
rep_smpl_screening <- function(data, BYr, rep, CI = 0.95) {
  # Step 1: Initialize storage for results (empty lists for each subpopulation)
  screening_full <- list()
  screening_female <- list()
  screening_male <- list()
  
  # Step 2: Loop over repetitions
  for (seed in 1:rep) {
    # Sample once for the full population
    smpl_full <- sample_SHG(data, BYr, seed)
    
    # Filter for male and female subpopulations (Sex: 1 = Female, 0 = Male)
    smpl_female <- smpl_full %>% filter(Sex == 1)
    smpl_male <- smpl_full %>% filter(Sex == 0)
    
    # Compute screening eligibility for each sample
    screening_full[[seed]] <- screening_eligible(smpl_full, dataset_type = "SHG")
    screening_female[[seed]] <- screening_eligible(smpl_female, dataset_type = "SHG")
    screening_male[[seed]] <- screening_eligible(smpl_male, dataset_type = "SHG")
  }
  
  # Step 3: Convert list of results into a single dataframe
  final_table <- bind_rows(
    aggregate_screening(screening_full, "Full Population", CI),
    aggregate_screening(screening_female, "Female Population", CI),
    aggregate_screening(screening_male, "Male Population", CI)
  )
  
  return(final_table)  # **Returns a single dataframe**
}
```


#format_shg
This function takes a vector of three SHG-derived percentage values (mean, lower bound, upper bound) and formats them as "mean (lower, upper)". This ensures a clear and standardized representation of SHG confidence intervals in tables.

```{r}
# Helper function to format SHG percentages
format_shg <- function(values, decimals = 1) {
  paste0(
    round(values[1], decimals), " (", 
    round(values[2], decimals), ", ", 
    round(values[3], decimals), ")"
  )
}
```


#compare_Data_SHG
This function compares smoking and screening eligibility statistics between SCCS data and SHG simulated results across multiple subpopulations. It outputs a table with SCCS values and SHG estimates formatted as “Mean (Lower, Upper)”. When diff = "On", the SHG values are reported as differences from the SCCS values. The function supports customization of the included subpopulations, optional percentage-based outputs, exclusion of ever smokers from denominators, and toggling of total population rows.

```{r}
compare_Data_SHG <- function(
  elig_Data_full, elig_Data_wh, elig_Data_bl,
  SHG_elig_full,  SHG_elig_wh,  SHG_elig_bl,
  total      = "Off",
  percentage = "Off",
  ever       = "Off",
  comparison = "All",
  diff       = "Off"
) {
  # 1. Subpop definitions
  comparison_map <- list(
    All   = c("Full","Female","Male","Black","Black Female","Black Male",
              "White","White Female","White Male"),
    FM    = c("Female","Male"),
    AFM   = c("Full","Female","Male"),
    WhFM  = c("White","White Female","White Male"),
    BlFM  = c("Black","Black Female","Black Male"),
    BlWh  = c("Black","White"),
    ABlWh = c("Full","Black","White"),
    MBW   = c("Male","Black Male","White Male"),
    FBW   = c("Female","Black Female","White Female")
  )
  selected_subpops <- comparison_map[[comparison]]
  
  # 2. Population‐to‐label maps
  subpop_map_data <- c(
    Full = "Full", Female = "Female", Male = "Male",
    Black = "Black", `Black Female` = "Black Female", `Black Male` = "Black Male",
    White = "White", `White Female` = "White Female", `White Male` = "White Male"
  )
  subpop_map_shg <- c(
    Full = "Full Population", Female = "Female Population", Male = "Male Population",
    Black = "Full Population", `Black Female` = "Female Population",
    `Black Male` = "Male Population",
    White = "Full Population", `White Female` = "Female Population",
    `White Male` = "Male Population"
  )
  
  # 3. Input lists
  data_df_list <- list(
    Full = elig_Data_full, Female = elig_Data_full, Male = elig_Data_full,
    Black = elig_Data_bl, `Black Female` = elig_Data_bl, `Black Male` = elig_Data_bl,
    White = elig_Data_wh, `White Female` = elig_Data_wh, `White Male` = elig_Data_wh
  )
  SHG_df_list <- list(
    Full = SHG_elig_full, Female = SHG_elig_full, Male = SHG_elig_full,
    Black = SHG_elig_bl, `Black Female` = SHG_elig_bl, `Black Male` = SHG_elig_bl,
    White = SHG_elig_wh, `White Female` = SHG_elig_wh, `White Male` = SHG_elig_wh
  )
  
  # 4. Measures + table skeleton
  measures <- c(
    "Total_Population","EverSmokers","CurrentSmokers","FormerSmokers",
    "PY20","PY30","SY20","SY30",
    "Eligible_2021","Eligible_2013","Eligible_SY20","Eligible_SY30"
  )
  table_cols <- unlist(lapply(selected_subpops, function(x) {
    c(paste0(x, "_Data"), paste0(x, "_SHG"))
  }))
  comparison_table <- setNames(
    as.data.frame(matrix(NA, length(measures), length(table_cols))),
    table_cols
  )
  rownames(comparison_table) <- measures
  
  # 5. Which measures use ever-smoker base?
  percent_by_ever_only <- c(
    "PY20","PY30","SY20","SY30",
    "Eligible_2021","Eligible_2013","Eligible_SY20","Eligible_SY30"
  )
  
  # 6. Fill in each subpop × measure
  for (subpop in selected_subpops) {
    data_row    <- data_df_list[[subpop]][
      data_df_list[[subpop]]$Population == subpop_map_data[subpop], ]
    SHG_row     <- SHG_df_list[[subpop]][
      SHG_df_list[[subpop]]$Population == subpop_map_shg[subpop], ]
    total_pop   <- data_row$Total_Population
    ever_smokers<- data_row$EverSmokers
    ever_prop   <- ever_smokers / 100
    total_prop  <- total_pop    / 100
    
    for (measure in measures) {
      # ————————————————
      # (A) SPECIAL-CASE Total_Population
      # ————————————————
      if (measure == "Total_Population") {
        Data_value  <- total_pop
        SHG_values  <- SHG_row[["Total_Population"]]
        data_dcmls  <- 0
        SHG_dcmls   <- 1   # or 0 if you prefer integer
      } else {
        # ————————————————
        # (B) Denominators & decimals for other measures
        # ————————————————
        if (percentage == "On"
            && ever == "On"
            && measure %in% percent_by_ever_only) {
          data_denum <- ever_prop
          SHG_denum  <- SHG_row$EverSmokers_Mean / 100
          data_dcmls <- 2; SHG_dcmls <- 2
        } else if (percentage == "On") {
          data_denum <- total_prop
          SHG_denum  <- total_prop
          data_dcmls <- 2; SHG_dcmls <- 2
        } else {
          data_denum <- 1
          SHG_denum  <- 1
          data_dcmls <- 0
          SHG_dcmls  <- 1
        }
        
        # Compute raw & SHG values
        Data_value <- round(data_row[[measure]] / data_denum, data_dcmls)
        mean_col   <- paste0(measure, "_Mean")
        lower_col  <- paste0(measure, "_Lower")
        upper_col  <- paste0(measure, "_Upper")
        vals       <- c(
          SHG_row[[mean_col]] / SHG_denum,
          SHG_row[[lower_col]] / SHG_denum,
          SHG_row[[upper_col]] / SHG_denum
        )
        if (diff == "On") vals <- vals - Data_value
        SHG_values <- format_shg(vals, decimals = SHG_dcmls)
      }
      
      # ————————————————
      # (C) Store into result
      # ————————————————
      comparison_table[measure, paste0(subpop, "_Data")] <- Data_value
      comparison_table[measure, paste0(subpop, "_SHG")]  <- SHG_values
    }
  }
  
  # 7. Optionally drop rows
  rows_to_remove <- character()
  if (total == "Off") rows_to_remove <- "Total_Population"
  # if (ever == "On") rows_to_remove <- c(rows_to_remove, "EverSmokers")
  if (length(rows_to_remove))
    comparison_table <- comparison_table[
      !rownames(comparison_table) %in% rows_to_remove, ]
  
  return(comparison_table)
}
```


#add_spacer_row
This helper inserts a blank HTML spacer row immediately after a named measure in a comparison table. It is used to visually separate logical groups of rows before the table is rendered or styled.

```{r}
add_spacer_row <- function(df, after_row_label) {
  idx <- which(df$Measure == after_row_label)
  if (length(idx) == 0) return(df)

  spacer <- df[1, ]
  spacer[1, ] <- "&nbsp;"  # fill all cells with spacer content
  df <- df %>% mutate(across(everything(), as.character))
  bind_rows(df[1:idx, ], spacer, df[(idx + 1):nrow(df), ])
}
```


#style_compare_table
This function formats the output of compare_Data_SHG into a visually enhanced HTML table using kableExtra. It applies color-coding to SHG values based on their relationship to the corresponding data values or zero (depending on diff = "On" or "Off"). It also adds hierarchical headers, bolds key columns and rows, and draws vertical separator lines between subpopulations. The function supports optional export to an HTML file and allows renaming of the data column label via the dataname argument.

```{r}
# Main styling function
style_compare_table <- function(df, difference = "Off", dataname = "Data", save_as_html = NULL) {
  styled_df <- df %>% tibble::rownames_to_column("Measure")
  col_names <- colnames(styled_df)[-1]
  subpops <- unique(gsub("_(Data|SHG)$", "", col_names))

  header_2 <- rep(c(dataname, "SHG"), length(subpops))
  header_1 <- setNames(rep(2, length(subpops)), subpops)

  # Colorize SHG entries
  for (col in col_names) {
    if (grepl("_SHG$", col)) {
      for (i in seq_len(nrow(styled_df))) {
        shg_val <- styled_df[[col]][i]
        data_colname <- gsub("_SHG$", "_Data", col)
        if (!data_colname %in% colnames(styled_df)) next
        data_val <- styled_df[[data_colname]][i]

        if (grepl("\\(", shg_val)) {
          parts <- str_match(shg_val, "([-+]?\\d*\\.?\\d+).*?\\(([-+]?\\d*\\.?\\d+), ([-+]?\\d*\\.?\\d+)\\)")
          mean <- as.numeric(parts[2])
          lower <- as.numeric(parts[3])
          upper <- as.numeric(parts[4])

          color <- if (difference == "On") {
            if (lower <= 0 && upper >= 0) "green"
            else if (upper < 0) "orange"
            else "blue"
          } else {
            if (lower <= data_val && upper >= data_val) "green"
            else if (upper < data_val) "orange"
            else "blue"
          }

          styled_df[[col]][i] <- paste0("<span style='color:", color, "'>", shg_val, "</span>")
        }
      }
    }
  }

  # Add section spacers
  styled_df <- add_spacer_row(styled_df, "FormerSmokers")  # after prevalence
  styled_df <- add_spacer_row(styled_df, "SY30")           # after pack-years

  # Format the table
  tbl <- kbl(
    styled_df,
    escape = FALSE,
    align = "c",
    col.names = c("Measure", header_2),
    booktabs = TRUE
  ) %>%
    add_header_above(c(" " = 1, header_1), bold = TRUE) %>%
    row_spec(which(styled_df$Measure == "&nbsp;"), extra_css = "height:20px;background-color:#f9f9f9;") %>%
    column_spec(1, bold = TRUE) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("condensed", "hover", "striped"))

  # Add vertical borders
  for (i in seq_along(subpops)) {
    data_col <- 2 * (i - 1) + 2
    shg_col  <- data_col + 1
    tbl <- tbl %>%
      column_spec(data_col, border_left = TRUE) %>%
      column_spec(shg_col, border_right = TRUE)
  }

  # Save if needed
  if (!is.null(save_as_html)) {
    save_kable(tbl, file = save_as_html)
  }

  return(tbl)
}
```


#style_compare_heatmap
Builds a custom HTML-based comparison table using raw HTML tags instead of kableExtra. Applies consistent styling and heatmap logic to highlight the closeness of SHG estimates to observed values. Suitable for more flexible rendering or embedding in HTML documents. Also supports optional export to a standalone HTML file.

```{r}
style_compare_heatmap <- function(
  df,
  difference     = "Off",
  rel_diff       = FALSE,
  save_as_html   = NULL,
  include_legend = TRUE,
  bins           = 20
) {
  # 1. Prepare dataframe and detect subpopulations
  styled_df <- df %>% tibble::rownames_to_column("Measure")
  col_names <- colnames(styled_df)[-1]
  subpops   <- unique(gsub("_(Data|SHG)$", "", col_names))
  
  # 2. Build the header row
  header_html <- paste0(
    "<tr>",
      "<th style='text-align:left;width:100px;border:1px solid #ccc;'>SHG vs SCCS</th>",
      paste0(
        "<th style='width:80px;border:1px solid #ccc;'>",
        subpops,
        "</th>",
        collapse = ""
      ),
    "</tr>"
  )
  
  # 3. Generate 'bins' shades of orange (light→dark) and 'bins' shades of blue (light→dark)
  orange_colors <- grDevices::colorRampPalette(c("#FFF8F0", "#E67300"))(bins)
  blue_colors   <- grDevices::colorRampPalette(c("#CCE5FF", "#004080"))(bins)
  
  # 4. Build each heatmap row
  row_html_list     <- list()
  insert_spacer_after <- c("FormerSmokers", "SY30")
  
  for (i in seq_len(nrow(styled_df))) {
    row_vals <- list()
    row_vals[[1]] <- sprintf(
      "<td style='font-weight:bold;text-align:left;width:100px;height:36px;border:1px solid #ccc;'>%s</td>",
      styled_df$Measure[i]
    )
    
    for (subpop in subpops) {
      data_val <- as.numeric(styled_df[[paste0(subpop, "_Data")]][i])
      shg_val  <- styled_df[[paste0(subpop, "_SHG")]][i]
      
      if (grepl("\\(", shg_val)) {
        parts        <- str_match(
          shg_val,
          "([-+]?\\d*\\.?\\d+).*?\\(([-+]?\\d*\\.?\\d+), ([-+]?\\d*\\.?\\d+)\\)"
        )
        mean_val     <- as.numeric(parts[2])
        lower        <- as.numeric(parts[3])
        upper        <- as.numeric(parts[4])
        sd_halfwidth <- (upper - lower) / 2
        
        # Compute direction & display_val
        if (difference == "On" && rel_diff) {
          rel_val     <- mean_val / data_val
          direction   <- mean_val
          distance    <- abs(mean_val)
          display_val <- sprintf("%+.1f%%", 100 * rel_val)
          
        } else if (difference == "Off" && rel_diff) {
          rel_val     <- (mean_val - data_val) / data_val
          direction   <- mean_val - data_val
          distance    <- abs(direction)
          display_val <- sprintf("%+.1f%%", 100 * rel_val)
          
        } else if (difference == "On") {
          direction   <- mean_val
          distance    <- abs(mean_val)
          display_val <- sprintf("%.2f", mean_val)
          
        } else {
          direction   <- mean_val - data_val
          distance    <- abs(direction)
          display_val <- sprintf("%.2f", data_val)
        }
        
        # Determine bin_idx: 0 if z ≤ 1; bin_idx = floor(z) for 1 < z < bins; bin_idx = bins if z ≥ bins
        z <- distance / sd_halfwidth
        if (z <= 1) {
          bin_idx <- 0
        } else if (z >= bins) {
          bin_idx <- bins
        } else {
          bin_idx <- floor(z)
        }
        
        # Pick fill_color & font_color
        if (bin_idx == 0) {
          fill_color <- "white"
          font_color <- "green"
          
        } else if (direction > 0) {
          # SHG > data → positive side uses blue_colors[bin_idx]
          fill_color <- blue_colors[bin_idx]
          if (bin_idx <= 2) {
            font_color <- "#000000"
          } else if (bin_idx >= bins - 1) {
            font_color <- "#FFFFFF"
          } else if (bin_idx <= floor(bins / 2)) {
            font_color <- "#222222"
          } else {
            font_color <- "#EEEEEE"
          }
        
        } else {
          # SHG < data → negative side uses orange_colors[bin_idx]
          fill_color <- orange_colors[bin_idx]
          if (bin_idx <= 2) {
            font_color <- "#000000"
          } else if (bin_idx >= bins - 1) {
            font_color <- "#FFFFFF"
          } else if (bin_idx <= floor(bins / 2)) {
            font_color <- "#222222"
          } else {
            font_color <- "#EEEEEE"
          }
        }
        
        cell_html <- sprintf(
          "<td style='background-color:%s;color:%s;text-align:center;font-weight:bold;width:80px;height:36px;border:1px solid #ccc;'>%s</td>",
          fill_color, font_color, display_val
        )
        
      } else {
        cell_html <- sprintf(
          "<td style='text-align:center;width:80px;height:36px;border:1px solid #ccc;'>%.2f</td>",
          data_val
        )
      }
      
      row_vals[[length(row_vals) + 1]] <- cell_html
    }
    
    row_html_list[[length(row_html_list) + 1]] <- 
      paste0("<tr>", paste(row_vals, collapse = ""), "</tr>")
    
    if (styled_df$Measure[i] %in% insert_spacer_after) {
      n_cols     <- length(subpops) + 1
      spacer_row <- paste0(
        "<tr>",
          paste(
            rep(
              "<td style='height:12px;border:none;background-color:transparent;'></td>",
              n_cols
            ),
            collapse = ""
          ),
        "</tr>"
      )
      row_html_list[[length(row_html_list) + 1]] <- spacer_row
    }
  }
  
  # 5. Build legend HTML if requested
  legend_html <- ""
  if (include_legend) {
    total_cells   <- 2 * bins + 1
    legend_colors <- c(
      rev(orange_colors),  # darkest→lightest so pale sits next to white
      "white",
      blue_colors         # lightest→darkest
    )
    
    # Initialize labels
    legend_labels <- rep("", total_cells)
    left_far    <- 1
    right_far   <- total_cells
    mid_index   <- bins + 1
    
    # Compute quarter markers
    q1 <- floor(bins / 4)
    q2 <- floor(bins / 2)
    q3 <- floor(3 * bins / 4)
    
    # Positions on orange side
    pos_o1 <- mid_index - q3
    pos_o2 <- mid_index - q2
    pos_o3 <- mid_index - q1
    
    # Positions on blue side
    pos_b1 <- mid_index + q1
    pos_b2 <- mid_index + q2
    pos_b3 <- mid_index + q3
    
    legend_labels[left_far]   <- "SHG < Data"
    legend_labels[pos_o1]     <- paste0(q3, " CI")
    legend_labels[pos_o2]     <- paste0(q2, " CI")
    legend_labels[pos_o3]     <- paste0(q1, " CI")
    legend_labels[mid_index]  <- "Agreement"
    legend_labels[pos_b1]     <- paste0(q1, " CI")
    legend_labels[pos_b2]     <- paste0(q2, " CI")
    legend_labels[pos_b3]     <- paste0(q3, " CI")
    legend_labels[right_far]  <- "SHG > Data"
    
    n_cols    <- length(subpops) + 1
    width_pct <- sprintf("%.2f%%", 100 / total_cells)
    
    # Build legend text cells; force nowrap on labeled positions
    text_cells <- sapply(seq_len(total_cells), function(i) {
      lbl <- legend_labels[i]
      if (i %in% c(left_far, pos_o1, pos_o2, pos_o3, mid_index, pos_b1, pos_b2, pos_b3, right_far)) {
        sprintf(
          "<td style='width:%s;text-align:center;font-size:8px;padding-top:2px;white-space:nowrap;'>%s</td>",
          width_pct, lbl
        )
      } else {
        sprintf(
          "<td style='width:%s;text-align:center;font-size:8px;padding-top:2px;'>%s</td>",
          width_pct, lbl
        )
      }
    })
    
    legend_html <- paste0(
      "<tr><td colspan='", n_cols,
         "' style='padding-top:15px;padding-bottom:4px;border-top:2px solid #ccc;'>",
        "<table style='width:100%;border-collapse:collapse;'>",
          "<tr>",
            paste0(
              "<td style='width:", width_pct,
              ";height:10px;border:1px solid #ccc;background-color:",
              legend_colors, ";'></td>",
              collapse = ""
            ),
          "</tr>",
          "<tr>",
            paste0(text_cells, collapse = ""),
          "</tr>",
        "</table>",
      "</td></tr>"
    )
  }
  
  # 6. Assemble full table HTML
  full_table <- paste0(
    "<table style='border-collapse:collapse;font-family:sans-serif;'>",
      header_html,
      paste0(row_html_list, collapse = "\n"),
      legend_html,
    "</table>"
  )
  
  if (!is.null(save_as_html)) writeLines(full_table, con = save_as_html)
  htmltools::HTML(full_table)
}
```


#plot_deviance_forest
This function generates a 33‑row forest plot of percent error for simulation outputs relative to observed data across All Races, Black, and White subpopulations, split by Overall, Female, and Male categories. You can optionally specify a symmetric x‑axis limit and confidence‑interval level for the legend.

```{r}
plot_deviance_forest <- function(df,
                                 title    = NULL,
                                 data = "Data",
                                 xlim     = 2.5,
                                 ci_level = 99,
                                 cohort_label = "SCCS") {
  name_map <- c(
    EverSmokers    = "Ever Smoked",
    CurrentSmokers = "Currently Smoke",
    FormerSmokers  = "Formerly Smoked",
    PY20           = "≥20 Pack Years",
    PY30           = "≥30 Pack Years",
    Eligible_2021  = "2021 USPSTF",
    Eligible_2013  = "2013 USPSTF",
    SY20           = "≥20-Yr Duration",
    SY30           = "≥30-Yr Duration",
    Eligible_SY20  = "USPSTF 20-Yr Duration",
    Eligible_SY30  = "USPSTF 30-Yr Duration"
  )

  race_colors <- c(
    "All Races" = "#c6dbef",
    "Black"     = "#4292c6",
    "White"     = "#084594"
  )

  ci_label <- paste0(ci_level, "% CI")

  df2 <- df %>% tibble::rownames_to_column("Measure")

  df_data <- df2 %>%
    pivot_longer(
      ends_with("_Data"),
      names_to = c("Subgroup", "Type"),
      names_pattern = "^(.*)_(Data)$",
      values_to = "SCCS"
    ) %>%
    select(Measure, Subgroup, SCCS)

  df_shg <- df2 %>%
    pivot_longer(
      ends_with("_SHG"),
      names_to = c("Subgroup", "Type"),
      names_pattern = "^(.*)_(SHG)$",
      values_to = "SHG_str"
    ) %>%
    select(Measure, Subgroup, SHG_str)

  df_long <- inner_join(df_data, df_shg, by = c("Measure", "Subgroup")) %>%
    mutate(
      SHG_mean = as.numeric(str_extract(SHG_str, "^[0-9.]+")),
      CI_low   = as.numeric(str_extract(SHG_str, "(?<=\\()[0-9.]+(?=, )")),
      CI_high  = as.numeric(str_extract(SHG_str, "(?<=, )[0-9.]+(?=\\))"))
    )

  df_plot <- df_long %>%
    mutate(
      D         = (SHG_mean - SCCS) / SCCS,
      D_low     = (CI_low   - SCCS) / SCCS,
      D_high    = (CI_high  - SCCS) / SCCS,
      Gender    = case_when(
        Subgroup %in% c("Full", "Black", "White") ~ "Overall",
        grepl("Female", Subgroup) ~ "Female",
        grepl("Male", Subgroup)   ~ "Male",
        TRUE ~ NA_character_
      ),
      Race = case_when(
        Subgroup %in% c("Full", "Female", "Male") ~ "All Races",
        grepl("Black", Subgroup) ~ "Black",
        grepl("White", Subgroup) ~ "White",
        TRUE ~ NA_character_
      ),
      BaseMeasure = str_remove(Measure, " Female$| Male$"),
      LabelBase = name_map[BaseMeasure],
      Category = paste0(LabelBase, if_else(Gender == "Overall", "", paste0(" ", Gender)))
    ) %>%
    filter(!is.na(Race) & !is.na(Gender))

  # Build a fixed row order with spacer rows separating prevalence, USPSTF, and duration metrics.
  base_groups <- names(name_map)
  group_labels <- name_map[base_groups]

  category_order <- character(0)
  for (i in seq_along(group_labels)) {
    category_order <- c(
      category_order,
      paste0(group_labels[i]),
      paste0(group_labels[i], " Female"),
      paste0(group_labels[i], " Male")
    )
    if (i == 3) category_order <- c(category_order, "__spacer1__")
    if (i == 7) category_order <- c(category_order, "__spacer2__")
  }

  # Spacer rows preserve visual grouping without affecting plotted estimates.
  spacer_rows <- tibble(
    Measure = NA, Subgroup = NA, SCCS = NA, SHG_mean = NA,
    CI_low = NA, CI_high = NA, D = NA, D_low = NA, D_high = NA,
    Gender = NA, Race = "All Races", BaseMeasure = NA,
    LabelBase = NA,
    Category = category_order[grepl("__spacer[0-9]+__", category_order)]
  )

  df_plot$Category <- as.character(df_plot$Category)
  df_plot <- bind_rows(df_plot, spacer_rows)
  df_plot$Category <- factor(df_plot$Category, levels = rev(category_order))

  # Spacer factor levels stay in the plot, but their y-axis labels are hidden.
  y_labels <- levels(df_plot$Category)
  y_labels[y_labels %in% c("__top_spacer__", "__spacer1__", "__spacer2__")] <- ""

  p <- ggplot(df_plot, aes(y = Category)) +
    geom_errorbarh(
      aes(xmin = D_low, xmax = D_high, color = Race, linetype = ci_label),
      height = 0.2, size = 0.8, na.rm = TRUE
    ) +
    geom_point(
      aes(x = D, color = Race, shape = Race),
      size = 3, na.rm = TRUE
    ) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "orange", size = 1.2) +
    annotate(
      "text",
      x = xlim * 0.98,
      y = tail(levels(df_plot$Category), 1),
      label = cohort_label,
      hjust = 1,
      vjust = 1.1,
      size = 3.8,
      fontface = "bold",
      color = "orange"
    ) +
    coord_cartesian(clip = "on") +
    scale_x_continuous(
      limits = c(-xlim, xlim),
      breaks = seq(-xlim, xlim, by = 1),
      labels = scales::percent_format(accuracy = 1),
      expand = c(0, 0)
    ) +
    scale_color_manual(
      name = "Subpopulation",
      values = race_colors,
      guide = guide_legend(order = 1)
    ) +
    scale_shape_manual(
      name = "Subpopulation",
      values = c("All Races" = 16, "Black" = 17, "White" = 15),
      guide = guide_legend(order = 1)
    ) +
    scale_linetype_manual(
      name = NULL,
      values = setNames("solid", ci_label),
      labels = ci_label,
      guide = guide_legend(order = 2)
    ) +
    scale_y_discrete(labels = y_labels, expand = expansion(mult = c(0.02, 0.02))) +
    labs(
      title = title,
      x = "Percent Error ((Simulated-Observed)/Observed)",
      y = NULL
    ) +
    theme_minimal() +
    theme(
      axis.text.y     = element_text(size = 8),
      legend.position = "bottom",
      legend.box      = "horizontal"
    )

  annotation_plot <- ggplot() +
    annotate(
      "segment", x = 0.2, xend = 2.5, y = 0, yend = 0,
      arrow = arrow(type = "closed", length = unit(0.15, "inches")),
      size = 0.6, color = "black"
    ) +
    annotate(
      "segment", x = -0.2, xend = -2.5, y = 0, yend = 0,
      arrow = arrow(type = "closed", length = unit(0.15, "inches")),
      size = 0.6, color = "black"
    ) +
    annotate(
      "text", x = 1.5, y = 0.025,
      label = "Overestimation", hjust = 0.5, vjust = 0, size = 3.5
    ) +
    annotate(
      "text", x = -1.5, y = 0.025,
      label = "Underestimation", hjust = 0.5, vjust = 0, size = 3.5
    ) +
    xlim(-2.5, 2.5) +
    ylim(-0.025, 0.1) +
    theme_void()

  return(annotation_plot / p + plot_layout(heights = c(0.1, 1)))
}
```


#compare_scatter_SCCS_SHG
This function generates a scatter plot comparing smoking intensity (cigarettes per day) and smoking duration (years) between the SCCS dataset and a sampled subset of the SHG dataset, incorporating jitter to improve visualization of overlapping points. It aligns the SCCS dataset’s sex encoding with SHG, extracts the birth year-gender distribution, and samples a single SHG dataset accordingly. The filtered ever-smoker data points are plotted with SCCS represented by orange triangles and SHG by blue circles. Users can adjust axis labels, the plot title, legend visibility, and the sampling coverage factor for SHG.

```{r}
compare_scatter_SCCS_SHG <- function(data_df, SHG_df, 
                                     plot_title = "Smoking Intensity vs. Duration", 
                                     x_title = "Cigarettes Per Day",
                                     y_title = "Smoking Duration (Years)",
                                     legend_on = TRUE,
                                     coverage = 1,
                                     seed = 42,
                                     threshold = "None") {
  
  # Step 1: Convert SCCS Sex encoding (M -> 0, F -> 1)
  data_df <- data_df %>%
    mutate(Sex = ifelse(Sex == "M", 0, 1))  # Now matches SHG format
  
  # Step 2: Extract birth year/gender composition from data_df
  byr_trgt <- sufftrgt(SHGdat2BY(data_df), coverage = coverage)
  
  # Step 3: Sample a single SHG dataset based on birth-year target
  sampled_SHG <- sample_SHG(SHG_df, byr_trgt, seed)
  
  # Step 4: Filter both datasets for Ever Smokers
  data_filtered <- data_df %>% filter(EverSmoker == 1)
  SHG_filtered <- sampled_SHG %>% filter(Eversmoker == 1)
  
  # Step 5: Create Scatter Plot with Jitter
  p <- ggplot() +
    geom_jitter(data = SHG_filtered, aes(y = Smoking_Years, x = CigsPerDay, shape = "SHG", color = "SHG"), 
                alpha = 0.15, size = 1, width = 0.5, height = 0.5) +  # SHG: Jittered Circles
    geom_jitter(data = data_filtered, aes(y = SmokingYears, x = CigsPerDay, shape = "SCCS", color = "SCCS"), 
                alpha = 0.15, size = 1, width = 0.5, height = 0.5) +  # SCCS: Jittered Circles
    scale_color_manual(
      values = c("SHG" = "blue", "SCCS" = "orange"),  # Correct color mapping
      name = "Dataset"
    ) +
    scale_shape_manual(
      values = c("SHG" = 16, "SCCS" = 16),  # SHG = Circle (16), SCCS = Circle (16)
      name = "Dataset"
    ) +
    theme_minimal() +
    coord_cartesian(xlim = c(0, 80), ylim = c(0, 70))  # Restrict CPD range
  
  # Step 6: Handle Optional Features
  
  # Add PY or SY threshold line
  if (threshold=="PY" | threshold == "Both"){
    p <- p + geom_function(
      fun = function(x) 400 / x,  # y = 400/x
      color = "darkred",
      linetype = "dashed",
      size = 1.5
    )
  } 
  if (threshold=="SY" | threshold=="Both"){
      p <- p + geom_function(
      fun = function(x) 20,  # y = 20
      color = "darkgreen",
      linetype = "dashed",
      size = 1.5
    )
  }
    
  # Set the title if provided, otherwise remove it
  if (!is.null(plot_title)) {
    p <- p + labs(title = plot_title)
  }
  
  # Set axis titles independently
  if (!is.null(x_title)) {
    p <- p + labs(x = x_title)
  } else {
    p <- p + labs(x = NULL)
  }
  
  if (!is.null(y_title)) {
    p <- p + labs(y = y_title)
  } else {
    p <- p + labs(y = NULL)
  }
  
  # Handle legend visibility
  if (!legend_on) {
    p <- p + theme(legend.position = "none")
  } else {
    p <- p + theme(legend.position = "bottom") +
      guides(color = guide_legend(override.aes = list(size = 3, alpha = 0.8)))
  }
  
  return(p)
}
```


#compare_scatter_SCCS_SHG_binned
Generates a scatter or binned count plot comparing SCCS and SHG smoking exposure data. Offers options for categorical binning, overlaying exposure thresholds, customizing titles, and legend display. Helps visually assess agreement between real and simulated data across key smoking dimensions.

```{r}
compare_scatter_SCCS_SHG_binned <- function(
  data_df, SHG_df,
  plot_title      = "Smoking Intensity vs. Duration",
  x_title         = "Cigarettes Per Day",
  y_title         = "Smoking Duration (Years)",
  legend_on       = TRUE,
  coverage        = 1,
  seed            = 42,
  threshold       = "None",
  shg_categorical = FALSE,
  arrowlabel      = FALSE,
  Ever            = TRUE
) {
  # Step 1: Convert SCCS Sex encoding
  data_df <- data_df %>% mutate(Sex = ifelse(Sex == "M", 0, 1))

  # Step 2: SHG data prep
  if (!shg_categorical) {
    byr_trgt    <- sufftrgt(SHGdat2BY(data_df), coverage = coverage)
    sampled_SHG <- sample_SHG(SHG_df, byr_trgt, seed)

    data_filtered <- data_df     %>% filter(EverSmoker  == 1)
    SHG_filtered  <- sampled_SHG %>% filter(Eversmoker == 1)

    p <- ggplot() +
      geom_jitter(
        data   = SHG_filtered,
        aes(x = CigsPerDay, y = Smoking_Years, color = "SHG"),
        alpha  = 0.15, size = 1, width = 0.5, height = 0.5
      ) +
      geom_jitter(
        data   = data_filtered,
        aes(x = CigsPerDay, y = SmokingYears, color = "SCCS"),
        alpha  = 0.15, size = 1, width = 0.5, height = 0.5
      )

    xrange <- c(0, 80)
    yrange <- c(0, 70)

  } else {
    # Step 3: Bin SCCS data into the same open-ended categories used for the bubble plots.
    data_binned <- data_df %>%
      filter(EverSmoker == 1) %>%
      mutate(
        CigsPerDay_cat = case_when(
          CigsPerDay <= 5  ~ 3,
          CigsPerDay <= 10 ~ 8,
          CigsPerDay <= 20 ~ 15,
          CigsPerDay <= 30 ~ 25,
          CigsPerDay >  30 ~ 35,
          TRUE             ~ NA_real_
        ),
        SmokingYears_cat = case_when(
          SmokingYears <= 10 ~ 5,
          SmokingYears <= 20 ~ 15,
          SmokingYears <= 30 ~ 25,
          SmokingYears <= 40 ~ 35,
          SmokingYears >  40 ~ 45,
          TRUE               ~ NA_real_
        )
      ) %>%
      count(CigsPerDay_cat, SmokingYears_cat, name = "SCCS_count") %>%
      mutate(color = "SCCS")

    # Step 4: Bin SHG data directly using the same category boundaries.
    SHG_binned <- SHG_df %>%
      filter(Eversmoker == 1) %>%
      mutate(
        CigsPerDay_cat = case_when(
          CigsPerDay <= 5  ~ 3,
          CigsPerDay <= 10 ~ 8,
          CigsPerDay <= 20 ~ 15,
          CigsPerDay <= 30 ~ 25,
          CigsPerDay >  30 ~ 35,
          TRUE             ~ NA_real_
        ),
        SmokingYears_cat = case_when(
          Smoking_Years <= 10 ~ 5,
          Smoking_Years <= 20 ~ 15,
          Smoking_Years <= 30 ~ 25,
          Smoking_Years <= 40 ~ 35,
          Smoking_Years >  40 ~ 45,
          TRUE                ~ NA_real_
        )
      ) %>%
      count(CigsPerDay_cat, SmokingYears_cat, name = "SHG_count") %>%
      mutate(SHG_count = SHG_count / 10, color = "SHG")

    # Step 5: Normalize bubble sizes to within-dataset shares when comparing ever-smoker distributions.
    if (Ever) {
      data_binned <- data_binned %>%
        mutate(SCCS_count = SCCS_count / sum(SCCS_count))

      SHG_binned <- SHG_binned %>%
        mutate(SHG_count = SHG_count / sum(SHG_count))
    }

    # Step 6: Plot binned points; point area reflects the binned share/count for each dataset.
    p <- ggplot() +
      geom_point(
        data = SHG_binned,
        aes(
          x = CigsPerDay_cat, y = SmokingYears_cat,
          size = SHG_count, color = color
        ),
        alpha = 0.6
      ) +
      geom_point(
        data = data_binned,
        aes(
          x = CigsPerDay_cat, y = SmokingYears_cat,
          size = SCCS_count, color = color
        ),
        alpha = 0.6
      ) +
      scale_size_area(max_size = 15, guide = FALSE)

    xrange <- c(0, 40)
    yrange <- c(0, 50)
  }

  # Step 7: Threshold overlays
  if (threshold %in% c("PY", "Both")) {
    p <- p +
      stat_function(
        aes(color = "20-Pack-Year Threshold"),
        fun      = function(x) 400 / x,
        size     = 1.2,
        xlim     = xrange,
        linetype = "dashed"
      )

    x_right_end <- max(xrange)
    y_right_end <- 400 / x_right_end
    x_top_end   <- 400 / max(yrange)
    y_top_end   <- max(yrange)
    dx <- (xrange[2] - xrange[1]) * 0.02
    dy <- (yrange[2] - yrange[1]) * 0.02

    p <- p +
      annotate("segment", x = x_right_end, xend = x_right_end,
               y = y_right_end + dy, yend = y_right_end + 5 * dy,
               arrow = grid::arrow(length = unit(0.025, "npc"), type = "closed"),
               color = "darkred", size = 1)

    if (arrowlabel) {
      p <- p +
        annotate("text", x = x_right_end,
                 y = y_right_end + 10 * dy + 5.5 * dy,
                 label = "\u2265 20-Pack-Yrs", color = "darkred",
                 size = 4, angle = 90, vjust = 0.5)
    }

    p <- p +
      annotate("segment", x = x_top_end + dx, xend = x_top_end + 5 * dx,
               y = y_top_end, yend = y_top_end,
               arrow = grid::arrow(length = unit(0.025, "npc"), type = "closed"),
               color = "darkred", size = 1)

    if (arrowlabel) {
      p <- p +
        annotate("text", x = x_top_end + 6 * dx, y = y_top_end,
                 label = "\u2265 20-Pack-Yrs", color = "darkred",
                 size = 4, hjust = 0)
    }
  }

  if (threshold %in% c("SY", "Both")) {
    sy_df <- data.frame(x = xrange[1], xend = xrange[2], y = 20, yend = 20)
    p <- p +
      geom_segment(
        data = sy_df,
        aes(x = x, xend = xend, y = y, yend = yend, color = "20-Year Duration Threshold"),
        size = 1.2, inherit.aes = FALSE, linetype = "dotted"
      )
  }

  # Step 8: Final styling
  if (shg_categorical) {
    # Major grid/tick locations match the displayed bin break labels in bubble mode.
    x_breaks <- c(0, 5, 10, 20, 30)
    y_breaks <- c(0, 10, 20, 30, 40)
  } else {
    x_breaks <- seq(xrange[1], xrange[2], 10)
    y_breaks <- seq(yrange[1], yrange[2], 10)
  }

  # Add a small amount of headroom on the upper bounds so threshold arrows fit without moving the origin.
  leeway <- c(0, 1)

  p <- p +
    # Draw the main axes explicitly at the origin while keeping the rest of the panel border off.
    geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
    geom_vline(xintercept = 0, color = "black", linewidth = 0.4) +
    scale_x_continuous(limits = xrange + leeway, breaks = x_breaks, expand = c(0, 0)) +
    scale_y_continuous(limits = yrange + leeway, breaks = y_breaks, expand = c(0, 0)) +
    scale_color_manual(
      name   = NULL,
      values = c(
        "SCCS" = "orange", "SHG" = "blue",
        "20-Pack-Year Threshold"     = "darkred",
        "20-Year Duration Threshold" = "darkgreen"
      ),
      breaks = c("SCCS", "SHG", "20-Pack-Year Threshold", "20-Year Duration Threshold"),
      labels = c(
        "SCCS (observed)",
        "SHG (simulated)",
        "20-Pack-Year Threshold",
        "20-Year Duration Threshold"
      ),
      drop   = FALSE
    ) +
    theme_minimal()

  if (!is.null(plot_title)) p <- p + labs(title = plot_title)
  if (!is.null(x_title))    p <- p + labs(x = x_title) else p <- p + labs(x = NULL)
  if (!is.null(y_title))    p <- p + labs(y = y_title) else p <- p + labs(y = NULL)

  if (!legend_on) {
    p <- p + theme(
      legend.position = "none",
      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank()
    )
  } else {
    # Keep only major grid lines so the grid aligns exactly with the chosen tick marks.
    p <- p + theme(
      legend.position = "bottom",
      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank()
    ) +
      guides(color = guide_legend(override.aes = list(size = 3, alpha = 0.8)))
  }

  return(p)
}
```


#compare_scatter_facet
This function generates a 2x2 grid of scatter plots comparing smoking intensity (cigarettes per day) and smoking duration (years) for four subpopulations: Black Female, Black Male, White Female, and White Male. It first converts the SCCS dataset’s sex encoding to match the SHG format, then filters the data by race and sex. Each subpopulation is plotted using the compare_scatter_SCCS_SHG function, ensuring appropriate axis labels and legend placement. Legends are shown only in the bottom row plots, while top row plots omit x-axis titles and legends for a cleaner presentation.

```{r}
compare_scatter_facet <- function(
  data_df, SHG_df, 
  coverage    = 1, 
  seed        = 42, 
  threshold   = "None",
  bubble      = TRUE,
  arrowlabel  = FALSE,
  facet_title = NULL
) {
  # Convert SCCS Sex encoding (M -> 0, F -> 1)
  data_df <- data_df %>% mutate(Sex = ifelse(Sex == "M", 0, 1))
  
  # Define the four SCCS subpopulations shown in the 2x2 panel layout.
  # SCCS sex is recoded above to match SHG conventions (F = 1, M = 0).
  subpops <- list(
    "Black Females" = data_df %>% filter(Sex == 1, RaceAnalysis == 2),
    "White Females" = data_df %>% filter(Sex == 1, RaceAnalysis == 1),
    "Black Males"   = data_df %>% filter(Sex == 0, RaceAnalysis == 2),
    "White Males"   = data_df %>% filter(Sex == 0, RaceAnalysis == 1)
  )
  
  # Generate four scatter plots
  plots <- list()
  
  plots[[1]] <- compare_scatter_SCCS_SHG_binned(
    subpops[["Black Females"]], SHG_df,
    plot_title      = "Black Females",
    x_title         = NULL,
    y_title         = "Smoking Duration (Years)",
    legend_on       = FALSE,
    seed            = seed,
    coverage        = coverage,
    threshold       = threshold,
    shg_categorical = bubble,
    arrowlabel      = arrowlabel
  )
  
  plots[[2]] <- compare_scatter_SCCS_SHG_binned(
    subpops[["White Females"]], SHG_df,
    plot_title      = "White Females",
    x_title         = NULL,
    y_title         = NULL,
    legend_on       = FALSE,
    seed            = seed,
    coverage        = coverage,
    threshold       = threshold,
    shg_categorical = bubble,
    arrowlabel      = FALSE
  )
  
  plots[[3]] <- compare_scatter_SCCS_SHG_binned(
    subpops[["Black Males"]], SHG_df,
    plot_title      = "Black Males",
    x_title         = "Cigarettes Per Day",
    y_title         = "Smoking Duration (Years)",
    legend_on       = FALSE,
    seed            = seed,
    coverage        = coverage,
    threshold       = threshold,
    shg_categorical = bubble,
    arrowlabel      = FALSE
  )
  
  plots[[4]] <- compare_scatter_SCCS_SHG_binned(
    subpops[["White Males"]], SHG_df,
    plot_title      = "White Males",
    x_title         = "Cigarettes Per Day",
    y_title         = NULL,
    legend_on       = FALSE,
    seed            = seed,
    coverage        = coverage,
    threshold       = threshold,
    shg_categorical = bubble,
    arrowlabel      = FALSE
  )
  
  # Build a single reference plot to extract the shared legend for the combined figure.
  legend_plot <- compare_scatter_SCCS_SHG_binned(
    subpops[["Black Males"]], SHG_df,
    plot_title      = NULL,
    x_title         = "Cigarettes Per Day",
    y_title         = "Smoking Duration (Years)",
    legend_on       = TRUE,
    seed            = seed,
    coverage        = coverage,
    threshold       = threshold,
    shg_categorical = bubble,
    arrowlabel      = arrowlabel
  )
  
  legend <- cowplot::get_legend(legend_plot)
  
  # Arrange plots into a 2x2 grid, then append the shared legend below.
  top_row <- plot_grid(plots[[1]], plots[[2]], ncol = 2, align = "hv")
  bottom_row <- plot_grid(plots[[3]], plots[[4]], ncol = 2, align = "hv")
  panel_grid <- plot_grid(top_row, bottom_row, ncol = 1, align = "hv")
  
  # Optional overall title
  if (!is.null(facet_title)) {
    title_grob <- ggdraw() + draw_label(facet_title, fontface = "bold", size = 14)
    final_plot <- plot_grid(title_grob, panel_grid, legend, ncol = 1, rel_heights = c(0.08, 1, 0.12))
  } else {
    final_plot <- plot_grid(panel_grid, legend, ncol = 1, rel_heights = c(1, 0.12))
  }
  
  return(final_plot)
}
```


## SCCS Data
# Load the SCCS data
This chunk of code prepares and processes smoking-related data from the SCCS dataset. It extracts relevant variables, cleans invalid/missing values, and calculates the SmokingYears variable based on smoking behavior. This chunk is foundational for analyzing smoking prevalence and its predictors.

```{r}
# Read SCCS input data and derive the cleaned smoking-history analysis dataset.
# `SCCS_sh_cleaned` is the main SCCS object used throughout the downstream workflow.

# Read the .dta file from the Southern Community Cohort Study (SCCS)
SCCS <- read_dta("SCCS_with_smoking_info_Lung_Breast_Colon.dta")

# View the data (uncomment to inspect the first few rows)
# head(SCCS)

# Define columns of interest
# Smoking-related variables
sh_cols <- c("EverSmoker", "AgeStartSmoking", "SmokeNow", "AgeQuitSmoking",
             "CigsPerDay", "SmokingStatus", "PackYears")

# Demographic and other key variables
info_cols <- c("Sex", "Enrollment_Age", "Enrollment_AgeMonths", "RaceAnalysis")

# Note: Check if additional columns like "RaceWhite" and "RaceBlack" exist if needed.

# Extract the relevant columns into a new dataframe
SCCS_sh <- SCCS[, c(info_cols, sh_cols)]

# Replace placeholder values (7777, 8888, 9999) with NA for missing or invalid data
SCCS_sh[SCCS_sh == 7777] <- NA
SCCS_sh[SCCS_sh == 8888] <- NA
SCCS_sh[SCCS_sh == 9999] <- NA

# Derive smoking duration from the smoking-history fields available in SCCS.
SCCS_sh <- SCCS_sh %>%
  mutate(
    SmokingYears = case_when(
      is.na(EverSmoker) ~ NA_real_, # If EverSmoker is missing, set SmokingYears to NA
      EverSmoker == 0 ~ 0,          # If the individual never smoked
      EverSmoker == 1 & SmokeNow == 1 ~ Enrollment_Age - AgeStartSmoking, # Currently smoking
      EverSmoker == 1 & SmokeNow == 0 ~ AgeQuitSmoking - AgeStartSmoking, # Quit smoking
      EverSmoker == 1 & is.na(SmokeNow) ~ Enrollment_Age - AgeStartSmoking, # Missing SmokeNow, assume still smoking
      TRUE ~ NA_real_ # Catch-all for any unexpected cases
    )
  )

# Keep only records with enough smoking-history detail to support the SCCS/SHG comparisons.
SCCS_sh_cleaned <- SCCS_sh %>%
  filter(
    !is.na(EverSmoker),
    !(EverSmoker == 1 & is.na(PackYears)),
    !(EverSmoker == 1 & SmokeNow == 0 & is.na(AgeQuitSmoking))
  )
```


## Visualizations: Distributions of CPD in SCCS
This chunk of code visualizes the distribution of cigarettes smoked per day (CigsPerDay) among Ever Smokers in the SCCS dataset. It separates the data into three groups: Full Population, White Population, and Black Population. For each group, the code calculates the number of individuals and overlays their density plots on a single graph. The title includes a summary of group counts, and the plot uses distinct colors and fills to differentiate between the groups, with the x-axis restricted to a plausible range for CigsPerDay.

```{r}
# Filter the data for EverSmoker == 1 and non-NA CigsPerDay
# Focus only on individuals who are Ever Smokers and have non-missing values for CigsPerDay.
filtered_data <- SCCS_sh %>%
  filter(EverSmoker == 1, !is.na(CigsPerDay))

# Create separate groups based on race or full population
# Full Population: Includes all Ever Smokers.
filtered_data_full <- filtered_data %>%
  mutate(RaceGroup = "Full Population")

# White Population: Ever Smokers identified as White (RaceAnalysis == 1).
filtered_data_white <- filtered_data %>%
  filter(RaceAnalysis == 1) %>%
  mutate(RaceGroup = "White")

# Black Population: Ever Smokers identified as Black (RaceAnalysis == 2).
filtered_data_black <- filtered_data %>%
  filter(RaceAnalysis == 2) %>%
  mutate(RaceGroup = "Black")

# Combine the three groups into a single dataframe
filtered_data_combined <- bind_rows(filtered_data_full, filtered_data_white, filtered_data_black)

# Calculate the number of individuals in each group for context
group_counts <- filtered_data_combined %>%
  group_by(RaceGroup) %>%
  summarise(Count = n()) %>%
  ungroup()

# Create a string summarizing group counts to display in the plot title
group_counts_text <- paste(
  group_counts$RaceGroup,
  group_counts$Count,
  sep = ": ",
  collapse = "; "
)

# Plot the distributions of CigsPerDay for each group
ggplot(filtered_data_combined, aes(x = CigsPerDay, color = RaceGroup, fill = RaceGroup)) +
  geom_density(alpha = 0.4) +  # Add density plots with some transparency
  coord_cartesian(xlim = c(0, 80)) +  # Focus on plausible CPD range
  labs(
    title = paste("Distributions of CigsPerDay Among Ever Smokers\n", group_counts_text),
    x = "Cigarettes Per Day",  # Label for x-axis
    y = "Density",  # Label for y-axis
    color = "Race Group",  # Legend for line colors
    fill = "Race Group"  # Legend for ribbon fills
  ) +
  theme_minimal()  # Clean and minimalistic plot theme
```


This chunk creates a compact baseline summary of the cleaned SCCS cohort, including demographic counts, smoking-history summaries, and the implied birth-year range. It serves as a quick reference table while checking the downstream SCCS/SHG comparisons.

```{r}
## ---------- Ever-smoker subset ----------
# `es` is used only for concise baseline summaries below.
es <- SCCS_sh_cleaned %>% 
  filter(EverSmoker == 1)

## ---------- Baseline-table summary ----------
# Baseline descriptive summary used as a quick reference for the cleaned SCCS cohort.
SCCS_demo <- SCCS_sh_cleaned %>% 
  summarise(
    # Demographic counts
    Male            = sum(Sex == "M", na.rm = TRUE),
    Female          = sum(Sex == "F", na.rm = TRUE),
    White           = sum(RaceAnalysis == 1, na.rm = TRUE),
    Black           = sum(RaceAnalysis == 2, na.rm = TRUE),
    Ever_Smoker     = nrow(es),
    Current_Smoker  = sum(SmokeNow == 1, na.rm = TRUE),
    Former_Smoker   = sum(EverSmoker == 1 & SmokeNow == 0, na.rm = TRUE),
    
    # Age (mean ± SD plus range)
    Mean_Age = mean(Enrollment_Age, na.rm = TRUE),
    SD_Age   = sd(  Enrollment_Age, na.rm = TRUE),
    Min_Age  = min(Enrollment_Age, na.rm = TRUE),
    Max_Age  = max(Enrollment_Age, na.rm = TRUE),
    
    # Smoking duration — both mean ± SD *and* median [IQR]
    Mean_SmokingYears   = mean(es$SmokingYears, na.rm = TRUE),
    SD_SmokingYears     = sd(  es$SmokingYears, na.rm = TRUE),
    Median_SmokingYears = median(es$SmokingYears, na.rm = TRUE),
    IQR_SmokingYears    = IQR(   es$SmokingYears, na.rm = TRUE),
    
    # Smoking intensity — median [IQR]
    Median_CigsPerDay = median(es$CigsPerDay, na.rm = TRUE),
    IQR_CigsPerDay    = IQR(   es$CigsPerDay, na.rm = TRUE),
    
    # Pack-years — median [IQR]
    Median_PackYears = median(es$PackYears, na.rm = TRUE),
    IQR_PackYears    = IQR(   es$PackYears, na.rm = TRUE),
    
    # Birth-year range (reference year 2005 for SCCS)
    Min_BY = 2005 - Max_Age,
    Max_BY = 2005 - Min_Age
  )

SCCS_demo

save.image("backup_WS.RData")
```


This chunk exports the cleaned SCCS smoking-history analysis dataset to CSV so it can be reviewed, shared, or inspected outside the RMarkdown workflow.

```{r}
# Export the cleaned SCCS analysis file for manual inspection and sharing outside R.
readr::write_csv(
  SCCS_sh_cleaned,
  "SCCS_sh_cleaned.csv",
  na = ""          # write missing values as empty cells
)
```



## Main code
# Creating target population sizes
This chunk calculates the birth year (YOB) for each individual in the cleaned SCCS dataset by subtracting enrollment age from the data collection year. It then summarizes the population by YOB and gender, producing three tables: one for the overall sample, one for White participants, and one for Black participants. Each table includes counts of males and females, grouped and sorted by year of birth.

```{r}
# Define the data collection year
data_collection_year <- 2005  # Assuming all individuals enrolled in this year

# Derive year of birth so SCCS targets can be aligned with SHG birth-year simulations.
SCCS_sh_cleaned <- SCCS_sh_cleaned %>%
  mutate(
    YOB = data_collection_year - Enrollment_Age  # Birth year is calculated as data collection year minus enrollment age
  )

# Build SCCS birth-year targets for the full cohort and for race-specific subcohorts.
birth_year_clean <- SCCS_sh_cleaned %>%
  group_by(YOB) %>%  # Group data by calculated birth year
  summarise(
    Females = sum(Sex == "F", na.rm = TRUE),  # Count females, excluding missing values
    Males = sum(Sex == "M", na.rm = TRUE)    # Count males, excluding missing values
  ) %>%
  arrange(YOB)  # Arrange the summary table by birth year

# Create the second data table: Population size by birth year for White population
birth_year_cl_white <- SCCS_sh_cleaned %>%
  filter(RaceAnalysis == 1) %>%  # Filter data for individuals classified as White
  group_by(YOB) %>%
  summarise(
    Females = sum(Sex == "F", na.rm = TRUE),  # Count females
    Males = sum(Sex == "M", na.rm = TRUE)    # Count males
  ) %>%
  arrange(YOB)  # Arrange the summary table by birth year

# Create the third data table: Population size by birth year for Black population
birth_year_cl_black <- SCCS_sh_cleaned %>%
  filter(RaceAnalysis == 2) %>%  # Filter data for individuals classified as Black
  group_by(YOB) %>%
  summarise(
    Females = sum(Sex == "F", na.rm = TRUE),  # Count females
    Males = sum(Sex == "M", na.rm = TRUE)    # Count males
  ) %>%
  arrange(YOB)  # Arrange the summary table by birth year
```


# Gender and Birth Year Composition Plot
This chunk combines the cleaned birth year distributions for the Full, Black, and White subpopulations into a single dataset, reshapes it into a tidy format separating Females and Males, and generates a faceted bar plot. The plot shows population size by year of birth and gender, with side-by-side panels for each subpopulation to enable visual comparison across groups.

```{r}
# Combine data for Full, Black, and White populations, adding a Subpopulation column
BYr_combined <- rbind(
  transform(birth_year_clean, Subpopulation = "Full"),  # Add "Full" subpopulation
  transform(birth_year_cl_black, Subpopulation = "Black"),  # Add "Black" subpopulation
  transform(birth_year_cl_white, Subpopulation = "White")   # Add "White" subpopulation
)

# Pivot the combined data into a tidy format
BYr_tidy <- pivot_longer(
  BYr_combined,
  cols = c(Females, Males),       # Columns to pivot (Gender categories)
  names_to = "Gender",            # New column for gender
  values_to = "Population"        # New column for population counts
)

# Create a faceted bar plot to visualize gender/age composition
ggplot(BYr_tidy, aes(x = YOB, y = Population, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +  # Create grouped bars for each gender
  facet_wrap(~ Subpopulation, ncol = 1) +           # Facet by subpopulation (one column)
  labs(
    title = "Gender/Age Composition by Subpopulation",  # Plot title
    x = "Year of Birth (YOB)",                         # x-axis label
    y = "Population",                                  # y-axis label
    fill = "Gender"                                    # Legend title
  ) +
  theme_minimal()                                      # Use a clean, minimal theme
```


## Combining and Exporting Birth-Year Summaries
This chunk merges three birth-year summary tables—one each for the full, White, and Black subpopulations in the cleaned dataset—into a single structured table. Columns are renamed for clarity, the data is aligned by year of birth (YOB), and the table is formatted using kableExtra for readability. It is then exported as a CSV file (table_cleaned_combined.csv) for potential inclusion in reports or further analysis.

```{r merge-birth-year-cleaned, echo=TRUE, message=FALSE, warning=FALSE}
# Merge the three birth-year targets into one exportable QA table.
table_cleaned_combined <- birth_year_clean %>%
  rename(Females_Full = Females, Males_Full = Males) %>%
  left_join(
    birth_year_cl_white %>% 
      rename(Females_White = Females, Males_White = Males),
    by = "YOB"
  ) %>%
  left_join(
    birth_year_cl_black %>% 
      rename(Females_Black = Females, Males_Black = Males),
    by = "YOB"
  )

# Create a styled table using kableExtra
BYR_table_cleaned <- table_cleaned_combined %>%
  kbl(
    caption = "Year of Birth (YOB) and Gender Counts by Subpopulation (Cleaned)",
    col.names = c("Year of Birth (YOB)", 
                  "Females (Full-Cleaned)", "Males (Full-Cleaned)",
                  "Females (White-Cleaned)", "Males (White-Cleaned)",
                  "Females (Black-Cleaned)", "Males (Black-Cleaned)"),
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )

# Export to CSV for further use in Excel or reporting
write.csv(table_cleaned_combined, "table_cleaned_combined.csv", row.names = FALSE)

save.image("backup_WS.RData")
```


# Creating simulated populations
This chunk of code creates simulated populations using a stochastic process (SHGbyr function) to match a target population size derived from the SCCS dataset (BYr_trgt). The process iteratively extends the simulation until the target size is reached, saving the final simulation results to a CSV file to avoid rerunning the time-intensive process. A pre-saved file can be reloaded for further analysis.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Create the SHG target population from the cleaned SCCS birth-year table.
# This is the target that the stochastic SHG runs are matched against.
BYr_trgt <- sufftrgt(birth_year_clean, 10)  # Adjust target size for the cleaned data


# !!!!! WARNING: THE FOLLOWING PROCESS IS TIME-CONSUMING AND CAN TAKE SEVERAL HOURS
# To avoid running this process, skip to the "Done!" section below.

# Run an initial SHG simulation, then extend it until birth-year coverage is sufficient.
sim_rslts <- SHGbyr(BYr_trgt, data_collection_year, seed = 40)  # Generate initial simulated data with seed 40
# Convert simulation results to birth year summary format
sim_byr <- SHGdat2BY(sim_rslts)  # Summarize simulated data by birth year
# Check population coverage by calculating shortages compared to the target
shrtge <- suff_shortage(sim_byr, BYr_trgt)  # Calculate shortages by birth year
initial_shrtge <- sum(shrtge[, 2:3])  # Total initial shortage across males and females

# Iteratively extend the simulation using a random seed of 30 until the target population size is reached
sim_comp_rslt <- SHGbyr_extend(sim_rslts, BYr_trgt, data_collection_year, seed = 30)  # Extend simulation with seed 30
# Convert updated simulation results to birth year summary format
sim_byr_ud <- SHGdat2BY(sim_comp_rslt)  # Summarize updated simulation data by birth year
# Check updated population coverage after extending the simulation
shrtge_ud <- suff_shortage(sim_byr_ud, BYr_trgt)  # Calculate updated shortages
final_shrtge <- sum(shrtge_ud[, 2:3])  # Total final shortage across males and females

# Save the final simulation results to a CSV file to avoid rerunning
write.csv(sim_comp_rslt, "sim_comp_rslt.csv", row.names = FALSE)  # Save results for future use

# !!!!! DONE! DO NOT RERUN THIS PROCESS UNLESS NECESSARY.
# Instead of rerunning, load the pre-saved simulation results below.

# Load pre-saved simulation results from a CSV file
# sim_comp_rslt <- read.csv("sim_comp_rslt.csv")  # Uncomment to load saved results

save.image("backup_WS.RData")
```


# Simulation Sampling and Extraction for Cleaned SCCS Data Redrawing for the cleaned SCCS data's sample size
This section redraws 10,000 simulated samples from the cleaned SCCS birth year distribution using a 99% confidence interval. It then extracts smoking prevalence and distribution measures for the full population, as well as male and female subgroups, based on the simulated SHG results. The final workspace is saved for backup.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Redraw repeated SCCS-sized samples from the SHG simulation for the full cleaned cohort.
smpl_rslts_clean <- rep_smpl_dist(sim_comp_rslt, birth_year_clean, 10000, CI = 0.99)

# Extract prevalence data (e.g., ever smokers, current smokers) from the cleaned sampled results
prvlnc_SHG_smpld_clean <- extr_prvlnc_smpld_rslts(smpl_rslts_clean)

# Extract distribution data for the full cleaned population
dist_SHG_smpld_clean <- extr_dist_smpld_rslts(smpl_rslts_clean)

# Extract distribution data for female and male subpopulations in cleaned data
dist_SHG_f_smpld_clean <- extr_dist_smpld_rslts(smpl_rslts_clean, population = "female")
dist_SHG_m_smpld_clean <- extr_dist_smpld_rslts(smpl_rslts_clean, population = "male")

save.image("backup_WS.RData")
```


This section performs repeated sampling (10,000 iterations, 99% CI) to match the cleaned birth year distribution of the White subpopulation. It extracts smoking prevalence and distribution measures for the full White group, and separately for White males and White females. The results are stored and the workspace is saved for backup.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Repeat the same redraw/extraction workflow for the cleaned White SCCS target.
smpl_rslts_wh_clean <- rep_smpl_dist(sim_comp_rslt, birth_year_cl_white, 10000, CI = 0.99)

# Extract prevalence data from the cleaned white population sample
prvlnc_SHG_wh_smpld_clean <- extr_prvlnc_smpld_rslts(smpl_rslts_wh_clean)
# Uncomment the following line to inspect the prevalence data
# head(prvlnc_SHG_wh_smpld_clean)

# Extract overall smoking distribution data from the cleaned white population sample
dist_SHG_wh_smpld_clean <- extr_dist_smpld_rslts(smpl_rslts_wh_clean)
# Uncomment the following line to inspect the distribution data
# head(dist_SHG_wh_smpld_clean)

# Extract smoking distribution data specifically for white females in cleaned data
dist_SHG_whf_smpld_clean <- extr_dist_smpld_rslts(smpl_rslts_wh_clean, population = "female")

# Extract smoking distribution data specifically for white males in cleaned data
dist_SHG_whm_smpld_clean <- extr_dist_smpld_rslts(smpl_rslts_wh_clean, population = "male")

save.image("backup_WS.RData")
```


This section generates 10,000 simulated samples (99% confidence interval) based on the cleaned birth year distribution of the Black subpopulation. It then extracts smoking prevalence and distribution data for the overall Black group, as well as separately for Black females and Black males. The session state is saved for reproducibility and backup.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Repeat the same redraw/extraction workflow for the cleaned Black SCCS target.
smpl_rslts_bl_clean <- rep_smpl_dist(sim_comp_rslt, birth_year_cl_black, 10000, CI = 0.99)

# Extract prevalence data from the cleaned black population sample
prvlnc_SHG_bl_smpld_clean <- extr_prvlnc_smpld_rslts(smpl_rslts_bl_clean)
# Uncomment the following line to inspect the prevalence data
# head(prvlnc_SHG_bl_smpld_clean)

# Extract overall smoking distribution data from the cleaned black population sample
dist_SHG_bl_smpld_clean <- extr_dist_smpld_rslts(smpl_rslts_bl_clean)
# Uncomment the following line to inspect the distribution data
# head(dist_SHG_bl_smpld_clean)

# Extract smoking distribution data specifically for black females in cleaned data
dist_SHG_blf_smpld_clean <- extr_dist_smpld_rslts(smpl_rslts_bl_clean, population = "female")

# Extract smoking distribution data specifically for black males in cleaned data
dist_SHG_blm_smpld_clean <- extr_dist_smpld_rslts(smpl_rslts_bl_clean, population = "male")

save.image("backup_WS.RData")
```


# Writing to xlsx
This batch of code creates an Excel workbook and stores 18 different SHG-related dataframes as separate sheets. These dataframes summarize smoking distribution results for different populations, including both raw and cleaned versions. The final workbook is saved as SHG_SCCS_smkdist.xlsx, making it easier to access and analyze these results later.
```{r}
# Export the cleaned SHG distribution summaries so they can be reviewed outside the Rmd.
wb_SHG_SCCS <- createWorkbook()

# List of the 18 dataframes sumamrizing SHG eligibility results for SCCS data
SHG_dataframes <- list(
#  "SHG" = dist_SHG_smpld, 
#  "SHG_m" = dist_SHG_m_smpld, 
#  "SHG_f" = dist_SHG_f_smpld,
  "SHG_cl" = dist_SHG_smpld_clean,
  "SHG_f_cl" = dist_SHG_f_smpld_clean,
  "SHG_m_cl" = dist_SHG_m_smpld_clean,
#  "SHG_bl" = dist_SHG_bl_smpld,
#  "SHG_blm" = dist_SHG_blm_smpld,
#  "SHG_blf" = dist_SHG_blf_smpld,
  "SHG_bl_cl" = dist_SHG_bl_smpld_clean,
  "SHG_blf_cl" = dist_SHG_blf_smpld_clean,
  "SHG_blm_cl" = dist_SHG_blm_smpld_clean,
#  "SHG_wh" = dist_SHG_wh_smpld, 
#  "SHG_whm" = dist_SHG_whm_smpld,
#  "SHG_whf" = dist_SHG_whf_smpld,
  "SHG_wh_cl" = dist_SHG_wh_smpld_clean,
  "SHG_whf_cl" = dist_SHG_whf_smpld_clean,
  "SHG_whm_cl" = dist_SHG_whm_smpld_clean 
)

# Loop through and add each dataframe to a separate sheet
for (name in names(SHG_dataframes)) {
  addWorksheet(wb_SHG_SCCS, name)
  writeData(wb_SHG_SCCS, name, SHG_dataframes[[name]])
}

# Save the workbook
saveWorkbook(wb_SHG_SCCS, "SHG_SCCS_smkdist.xlsx", overwrite = TRUE)
```



# Distribution Plots

# The graphs for the cleaned data and simulation results
This chunk produces side-by-side stacked plots comparing smoking history distributions between the cleaned SCCS dataset and SHG simulation results for White and Black populations. It uses a 99% confidence interval for both groups and leverages the stacked_side_plots function. The final figure is exported as a high-resolution PNG image (800x1000 pixels) suitable for presentations or reports.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Compare SHG and SCCS distributions side-by-side for the two race groups.
SCCS_BW_cl <- stacked_side_plots(
  dist_SHG_bl_smpld_clean,                  # Cleaned black population sampled distribution
  dist_SHG_wh_smpld_clean,                  # Cleaned white population sampled distribution
  SCCS_sh_cleaned[SCCS_sh_cleaned$RaceAnalysis == 2,], # Cleaned SCCS data for black population
  SCCS_sh_cleaned[SCCS_sh_cleaned$RaceAnalysis == 1,], # Cleaned SCCS data for white population
  maintitle = "Smoking History Distributions: SCCS vs. SHG",
  column_titles = c("Black Population", "White Population"), # Updated titles
  CI = 99  # Confidence interval for the plots
)

SCCS_BW_cl

# Save the plot as PNG with 800x1000 pixel dimensions
ggsave(
  filename = "SCCS_BW_cl.PNG", 
  plot = SCCS_BW_cl,
  width = 800 / 100,  # Convert pixels to inches (assuming 300 dpi)
  height = 1000 / 100, 
  dpi = 300
)
```


This code generates a side-by-side visualization of smoking history distributions for male and female populations, comparing cleaned SCCS data to SHG simulation results. The plots incorporate 99% confidence intervals and are exported as a high-resolution PNG image (800x1000 pixels) for use in reports or presentations.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Compare SHG and SCCS distributions side-by-side for the two sex groups.
SCCS_FM_cl <- stacked_side_plots(
  dist_SHG_f_smpld_clean,                  # Cleaned female population sampled distribution
  dist_SHG_m_smpld_clean,                  # Cleaned male population sampled distribution
  SCCS_sh_cleaned[SCCS_sh_cleaned$Sex == "F",], # Cleaned SCCS data for female population
  SCCS_sh_cleaned[SCCS_sh_cleaned$Sex == "M",], # Cleaned SCCS data for male population
  maintitle = "Smoking History Distributions: SCCS vs. SHG",
  column_titles = c("Female Population", "Male Population"), # Updated titles
  CI = 99  # Confidence interval for the plots
)

SCCS_FM_cl

# Save the plot as PNG with 800x1000 pixel dimensions
ggsave(
  filename = "SCCS_FM_cl.PNG", 
  plot = SCCS_FM_cl,
  width = 800 / 100,  # Convert pixels to inches (assuming 300 dpi)
  height = 1000 / 100, 
  dpi = 300
)
```


This chunk compares smoking history distributions between cleaned SCCS data and SHG simulation outputs for white and black male subgroups. It generates a side-by-side bar plot with 99% confidence intervals and labeled columns, then saves the figure as a high-resolution PNG file (800x1000 pixels) for reporting or presentation use.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Compare SHG and SCCS distributions within the male race-stratified subgroups.
SCCS_BWM_cl <- stacked_side_plots(
  dist_SHG_blm_smpld_clean,                                         # Cleaned black male population sampled distribution
  dist_SHG_whm_smpld_clean,                                         # Cleaned white male population sampled distribution
  SCCS_sh_cleaned[SCCS_sh_cleaned$Sex == "M" & SCCS_sh_cleaned$RaceAnalysis == 2, ],  # Cleaned SCCS data for black male population
  SCCS_sh_cleaned[SCCS_sh_cleaned$Sex == "M" & SCCS_sh_cleaned$RaceAnalysis == 1, ],  # Cleaned SCCS data for white male population
  maintitle = "Smoking History Distributions: SCCS vs. SHG",
  column_titles = c("Black Male Population", "White Male Population"), # Updated titles
  CI = 99  # Confidence interval for the plots
)

SCCS_BWM_cl

# Save the plot as PNG with 800x1000 pixel dimensions
ggsave(
  filename = "SCCS_BWM_cl.PNG", 
  plot = SCCS_BWM_cl,
  width = 800 / 100,  # Convert pixels to inches (assuming 300 dpi)
  height = 1000 / 100, 
  dpi = 300
)
```


This section visualizes and compares the smoking history distributions of white and black female populations using cleaned SCCS data and SHG simulation results. It uses stacked_side_plots() to create a faceted plot with 99% confidence intervals, clearly labeling each subgroup. The plot is then exported as a high-resolution PNG image for reporting or presentation (800x1000 pixels).

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Compare smoking distributions for white and black female populations using cleaned data
SCCS_BWF_cl <- stacked_side_plots(
  dist_SHG_blf_smpld_clean,                                         # Cleaned black female population sampled distribution
  dist_SHG_whf_smpld_clean,                                         # Cleaned white female population sampled distribution
  SCCS_sh_cleaned[SCCS_sh_cleaned$Sex == "F" & SCCS_sh_cleaned$RaceAnalysis == 2, ],  # Cleaned SCCS data for black female population
  SCCS_sh_cleaned[SCCS_sh_cleaned$Sex == "F" & SCCS_sh_cleaned$RaceAnalysis == 1, ],  # Cleaned SCCS data for white female population
  maintitle = "Smoking History Distributions: SCCS vs. SHG",
  column_titles = c("Black Female Population", "White Female Population"), # Updated titles
  CI = 99  # Confidence interval for the plots
)

SCCS_BWF_cl

# Save the plot as PNG with 800x1000 pixel dimensions
ggsave(
  filename = "SCCS_BWF_cl.PNG", 
  plot = SCCS_BWF_cl,
  width = 800 / 100,  # Convert pixels to inches (assuming 300 dpi)
  height = 1000 / 100, 
  dpi = 300
)
```


This command saves the entire current R workspace—including all variables, data, and loaded objects—into a file named backup_WS.RData. This allows you to reload your work later using load("backup_WS.RData").

```{r}
save.image("backup_WS.RData")
```






This chunk combines the race-only, female race-stratified, and male race-stratified distribution panels into one vertically stacked figure with a shared y-axis label. It is the more presentation-oriented summary figure built from the cleaned SCCS and SHG distribution plots.

```{r}
# First subplot: keep title, no legend
SCCS_BW <- stacked_plots(
  dist_SHG_bl_smpld_clean,
  dist_SHG_wh_smpld_clean,
  SCCS_sh_cleaned[SCCS_sh_cleaned$RaceAnalysis == 2,],
  SCCS_sh_cleaned[SCCS_sh_cleaned$RaceAnalysis == 1,],
  maintitle = "Smoking History Distributions: SCCS vs. SHG",
  row_titles = c("Black Adults", "White Adults"),
  CI = 99,
  include_title = FALSE,
  include_legend = FALSE  # suppress here
)

# Second: no title, no legend
SCCS_BWF <- stacked_plots(
  dist_SHG_blf_smpld_clean,                                         # Cleaned black female population sampled distribution
  dist_SHG_whf_smpld_clean,                                         # Cleaned white female population sampled distribution
  SCCS_sh_cleaned[SCCS_sh_cleaned$Sex == "F" & SCCS_sh_cleaned$RaceAnalysis == 2, ],  # Cleaned SCCS data for black female population
  SCCS_sh_cleaned[SCCS_sh_cleaned$Sex == "F" & SCCS_sh_cleaned$RaceAnalysis == 1, ],  # Cleaned SCCS data for white female population
  row_titles = c("Black Females", "White Females"),
  CI = 99,
  include_title = FALSE,
  include_legend = FALSE
)
# Third: no title, keep legend
SCCS_BWM <- stacked_plots(
  dist_SHG_blm_smpld_clean,
  dist_SHG_whm_smpld_clean,
  SCCS_sh_cleaned[SCCS_sh_cleaned$Sex == "M" & SCCS_sh_cleaned$RaceAnalysis == 2, ],
  SCCS_sh_cleaned[SCCS_sh_cleaned$Sex == "M" & SCCS_sh_cleaned$RaceAnalysis == 1, ],
  row_titles = c("Black Males", "White Males"),
  CI = 99,
  include_title = FALSE,
  include_legend = TRUE  # only this one includes legend
)

final_figure <- SCCS_BW /
                plot_spacer() /
                SCCS_BWF /
                plot_spacer() /
                SCCS_BWM +
  plot_layout(ncol = 1, heights = c(1, 0.1, 1, 0.1, 1.1))

y_axis_label <- ggdraw() +
  draw_label("Density", angle = 90, hjust = 0.5, size = 20)

# Combine into full layout with left-side label
final_figure_labeled <- plot_grid(
  y_axis_label,
  final_figure,
  ncol = 2,
  rel_widths = c(0.04, 1)
)

ggsave(
  filename = "SCCS_Combined_Plots_HL.png",
  plot = final_figure_labeled,
  width = 1500 / 100,  # 2400 px width at 300 dpi
  height = 2000 / 100, # 1000 px height
  dpi = 300,
  bg = "white"
)
```

This chunk produces a simplified Black-versus-White distribution figure only, with a shared y-axis label and a single legend. It is useful when the full multi-panel summary figure is more detail than needed.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
SCCS_BW_only <- stacked_plots(
  dist_SHG_bl_smpld_clean,
  dist_SHG_wh_smpld_clean,
  SCCS_sh_cleaned[SCCS_sh_cleaned$RaceAnalysis == 2, ],
  SCCS_sh_cleaned[SCCS_sh_cleaned$RaceAnalysis == 1, ],
  maintitle = "Smoking History Distributions: SCCS vs. SHG",
  row_titles = c("Black Adults", "White Adults"),
  CI = 99,
  include_title = FALSE,
  include_legend = TRUE
)

y_axis_label_bw <- ggdraw() +
  draw_label("Density", angle = 90, hjust = 0.5, size = 20)

SCCS_BW_only_labeled <- plot_grid(
  y_axis_label_bw,
  SCCS_BW_only,
  ncol = 2,
  rel_widths = c(0.04, 1)
)

SCCS_BW_only_labeled

ggsave(
  filename = "SCCS_BW.png",
  plot = SCCS_BW_only_labeled,
  width = 1500 / 100,
  height = 800 / 100,
  dpi = 300,
  bg = "white"
)
```

# Simulating Screening Eligibility Based on Cleaned Demographics
This code estimates lung cancer screening eligibility for the full, White, and Black subpopulations using the cleaned SCCS birth year distributions. The rep_smpl_screening() function runs 10,000 simulations per group to compute eligibility rates along with 99% confidence intervals. The results—stored in smpl_rslts_eligible_cl, smpl_rslts_elig_wh_cl, and smpl_rslts_elig_bl_cl—are saved for downstream analysis and visualization.

```{r}
# Repeated screening redraws use the same cleaned SCCS birth-year targets as the distribution redraws.
smpl_rslts_eligible_cl <- 
  rep_smpl_screening(sim_comp_rslt, birth_year_clean, 10000, CI = 0.99)

smpl_rslts_elig_wh_cl <- 
  rep_smpl_screening(sim_comp_rslt, birth_year_cl_white, 10000, CI = 0.99)

smpl_rslts_elig_bl_cl <- 
  rep_smpl_screening(sim_comp_rslt, birth_year_cl_black, 10000, CI = 0.99)

save.image("backup_WS.RData")
```


To compute lung cancer screening eligibility among the SCCS participants, we apply the screening_eligible() function to both the full population and race-gender subgroups within the cleaned SCCS dataset. This includes the full, male, and female groups, as well as stratified subsets for White and Black populations by sex. For each group, the results are tagged with a corresponding Population label to aid in later comparisons and visualizations. The outputs are organized into three separate combined data frames: one for the full SCCS population (elig_sccs_full_combined), one for the White subpopulations (elig_sccs_wh_combined), and one for the Black subpopulations (elig_sccs_bl_combined).

```{r}
# Assemble observed SCCS screening-eligibility summaries for the same subgroups used in SHG outputs.
# Keeping `Population` as the first column makes downstream comparison tables easier to read.

# full population
elig_sccs_full <- screening_eligible(SCCS_sh_cleaned, dataset_type = "SCCS")
elig_sccs_m <- screening_eligible(SCCS_sh_cleaned %>% filter(Sex == "M"), dataset_type = "SCCS")
elig_sccs_f <- screening_eligible(SCCS_sh_cleaned %>% filter(Sex == "F"), dataset_type = "SCCS")

elig_sccs_full_combined <- bind_rows(
  elig_sccs_full %>% mutate(Population = "Full"),
  elig_sccs_m %>% mutate(Population = "Male"),
  elig_sccs_f %>% mutate(Population = "Female")
) %>%
  select(Population, everything())  # Ensure "Population" is the first column

# white population
elig_sccs_wh <- screening_eligible(SCCS_sh_cleaned %>% filter(RaceAnalysis == 1), dataset_type = "SCCS")
elig_sccs_whm <- screening_eligible(SCCS_sh_cleaned %>% filter(RaceAnalysis == 1 & Sex == "M"), dataset_type = "SCCS")
elig_sccs_whf <- screening_eligible(SCCS_sh_cleaned %>% filter(RaceAnalysis == 1 & Sex == "F"), dataset_type = "SCCS")

# Merge into a single dataframe
elig_sccs_wh_combined <- bind_rows(
  elig_sccs_wh %>% mutate(Population = "White"),
  elig_sccs_whm %>% mutate(Population = "White Male"),
  elig_sccs_whf %>% mutate(Population = "White Female")
) %>%
  select(Population, everything())  # Ensure "Population" is the first column

# black population
elig_sccs_bl <- screening_eligible(SCCS_sh_cleaned %>% filter(RaceAnalysis == 2), dataset_type = "SCCS")
elig_sccs_blm <- screening_eligible(SCCS_sh_cleaned %>% filter(RaceAnalysis == 2 & Sex == "M"), dataset_type = "SCCS")
elig_sccs_blf <- screening_eligible(SCCS_sh_cleaned %>% filter(RaceAnalysis == 2 & Sex == "F"), dataset_type = "SCCS")

# Merge into a single dataframe
elig_sccs_bl_combined <- bind_rows(
  elig_sccs_bl %>% mutate(Population = "Black"),
  elig_sccs_blm %>% mutate(Population = "Black Male"),
  elig_sccs_blf %>% mutate(Population = "Black Female")
) %>%
  select(Population, everything())  # Ensure "Population" is the first column
```


# Constructing an Eligibility Comparison Table
This section generates a summary table comparing lung cancer screening eligibility estimates from the cleaned SCCS data and corresponding SHG simulations. The compare_Data_SHG() function is applied to full, White, and Black subpopulations using absolute counts rather than percentages. The resulting table, comp_table_cl_raw, highlights differences across key measures and is exported as eligibility_table_cl_raw.csv for use in reporting or further analysis.

```{r}
# Raw-count comparison table used to sanity-check simulated totals before moving to percentages.
comp_table_cl_raw <- compare_Data_SHG(elig_sccs_full_combined,
                               elig_sccs_wh_combined,
                               elig_sccs_bl_combined,
                               smpl_rslts_eligible_cl, 
                               smpl_rslts_elig_wh_cl, 
                               smpl_rslts_elig_bl_cl,
                               total = "On", 
                               percentage = "Off",
                               comparison = "All")

style_compare_table(comp_table_cl_raw, diff = "Off", dataname = "SCCS")

# Exports comp_table to a CSV file
# write.csv(comp_table_cl_raw, "eligibility_table_cl_raw.csv", row.names = TRUE)
```


This section generates a cleaned eligibility comparison table where all estimates are expressed as percentages of the total population. Using the compare_Data_SHG() function, SCCS data and SHG simulation outputs—each based on cleaned birth year distributions—are compared across all major subpopulations. The table includes the total population and excludes ever-smoker-based normalization. The resulting comparison, comp_table_cl, is saved as a CSV file named eligibility_table_cl.csv for downstream analysis and presentation.

```{r}
# Percent-of-total comparison table for the cleaned SCCS targets.
comp_table_cl <- compare_Data_SHG(elig_sccs_full_combined,
                               elig_sccs_wh_combined,
                               elig_sccs_bl_combined,
                               smpl_rslts_eligible_cl, 
                               smpl_rslts_elig_wh_cl, 
                               smpl_rslts_elig_bl_cl,
                               total = "On", 
                               percentage = "On",
                               ever = "Off",
                               comparison = "All")

# Exports comp_table to a CSV file
#write.csv(comp_table_cl, "eligibility_table_cl.csv", row.names = TRUE)

style_compare_table(comp_table_cl, diff = "Off", dataname = "SCCS")
```


This code constructs a comparison table where all eligibility rates are normalized as percentages of the ever-smoking population. The function compare_Data_SHG() is applied using cleaned SCCS and SHG simulation datasets for full, white, and Black subpopulations. With the ever = "On" flag, all percentages are relative to those who have ever smoked, rather than the total population. The generated table, comp_table_ev, is exported as a CSV file named eligibility_table_ev.csv for further reporting or visualization.

```{r}
# Percent-of-ever-smoker comparison table, used when eligibility is normalized to the smoking population.
comp_table_ev <- compare_Data_SHG(elig_sccs_full_combined,
                               elig_sccs_wh_combined,
                               elig_sccs_bl_combined,
                               smpl_rslts_eligible_cl, 
                               smpl_rslts_elig_wh_cl, 
                               smpl_rslts_elig_bl_cl,
                               total = "Off",
                               comparison = "All", 
                               percentage = "On",
                               ever = "On")

# Exports comp_table to a CSV file
# write.csv(comp_table_ev, "eligibility_table_ev.csv", row.names = TRUE)

style_compare_table(comp_table_ev, diff = "Off", dataname = "SCCS")
```


This section generates a comparison table of eligibility outcomes, where all values represent the difference between SHG simulation results and SCCS data, expressed as percentages of the ever-smoking population. By setting diff = "On", the table highlights discrepancies between observed and modeled values. The total population row is excluded (total = "Off") to focus on subgroup differences. The resulting table, comp_table_ev_diff, is exported as a CSV file named eligibility_table_ev_diff.csv for subsequent use in analysis or reporting.

```{r}
# Difference table on the ever-smoker scale.
comp_table_ev_diff <- compare_Data_SHG(elig_sccs_full_combined,
                               elig_sccs_wh_combined,
                               elig_sccs_bl_combined,
                               smpl_rslts_eligible_cl, 
                               smpl_rslts_elig_wh_cl, 
                               smpl_rslts_elig_bl_cl,
                               total = "Off",
                               comparison = "All", 
                               percentage = "On",
                               ever = "On",
                               diff = "On")

# Exports comp_table to a CSV file
# write.csv(comp_table_ev_diff, "eligibility_table_ev_diff.csv", row.names = TRUE)

style_compare_table(comp_table_ev_diff, diff = "On", dataname = "SCCS")
```


This block constructs a comparison table displaying the percentage-point differences between SHG simulation estimates and SCCS data, using the total population as the denominator. The diff = "On" setting calculates and presents these differences, while total = "Off" omits the overall population row. The table covers all specified subpopulations and is saved as a CSV file titled eligibility_table_diff.csv.

```{r}
# Difference table on the total-population scale.
comp_table_diff <- compare_Data_SHG(elig_sccs_full_combined,
                               elig_sccs_wh_combined,
                               elig_sccs_bl_combined,
                               smpl_rslts_eligible_cl, 
                               smpl_rslts_elig_wh_cl, 
                               smpl_rslts_elig_bl_cl,
                               total = "Off",
                               comparison = "All", 
                               percentage = "On",
                               ever = "Off",
                               diff = "On")

# Exports comp_table to a CSV file
# write.csv(comp_table_diff, "eligibility_table_diff.csv", row.names = TRUE)
style_compare_table(comp_table_diff, diff = "On", dataname = "SCCS")
```


This line saves the entire R workspace, including all current objects (data, functions, results), to a file named backup_WS.RData. This allows for easy restoration of the session later without needing to re-run previous code.

```{r}
save.image("backup_WS.RData")
```


This code block creates a styled comparison table by applying the style_compare_table() function to the comp_table_diff data frame. The function is called with the diff argument set to "On", which enables color-coding of the SHG estimates based on whether the simulated values differ significantly from the SCCS data. The dataname parameter is set to "SCCS", which ensures that the column headers labeled “Data” in the default output are now labeled “SCCS” to reflect the source of the data more accurately. The styled table is saved as an HTML file named "comp_table.html", allowing the formatted output to be preserved for use in external documents such as Word or PowerPoint. Finally, the last line displays the styled table within the RMarkdown document or R console, providing an immediate visual of the formatted results.

```{r}
# Save a formatted HTML version of the total-population difference table.
styled_table <- style_compare_table(comp_table_diff, diff = "On", 
                                    save_as_html = "comp_table.html",
                                    dataname = "SCCS")
styled_table
```


This code generates a styled comparison table that highlights differences between SCCS data and SHG simulation results for ever smokers. By passing the comp_table_ev_diff data frame into the style_compare_table() function with diff = "On", the table applies conditional formatting: SHG values are color-coded based on whether their confidence intervals indicate meaningful differences from the SCCS values. The dataname argument is set to "SCCS", so that columns originally labeled “Data” now reflect the actual data source. The resulting table is saved as an HTML file named "comp_ev_table.html", which retains its formatting for easy integration into reports or presentations. The final line ensures the styled table is also displayed in the output for immediate review.

```{r}
# Save a formatted HTML version of the ever-smoker difference table.
styled_table_ev <- style_compare_table(comp_table_ev_diff, diff = "On", 
                                    save_as_html = "comp_ev_table.html",
                                    dataname = "SCCS")
styled_table_ev
```


This line saves the entire R workspace, including all current objects (data, functions, results), to a file named backup_WS.RData. This allows for easy restoration of the session later without needing to re-run previous code.

```{r}
save.image("backup_WS.RData")
```


This chunk saves the styled total-population difference table directly to HTML without printing the returned object. It is a convenience export for slides, manuscripts, or external review.

```{r}
style_compare_table(comp_table_diff, diff = "On", 
                                    save_as_html = "comp_table.html",
                                    dataname = "SCCS")
```


This chunk exports the comparison table in heatmap-table form, keeping the table layout while applying the heatmap-style coloring used elsewhere in the report.

```{r}
style_compare_heatmap_table(comp_table_diff, difference="On", save_as_html = "comp_heatmap_table.html")
```


This chunk builds the no-difference comparison object used by the main forest plot and the corresponding heatmap export. It keeps SCCS values and SHG estimates on their original scales rather than converting them to differences.

```{r}
# Build the no-difference comparison object used by the forest plot and heatmap views.
comp_hm <- compare_Data_SHG(elig_sccs_full_combined,
                               elig_sccs_wh_combined,
                               elig_sccs_bl_combined,
                               smpl_rslts_eligible_cl, 
                               smpl_rslts_elig_wh_cl, 
                               smpl_rslts_elig_bl_cl,
                               total = "Off",
                               comparison = "All", 
                               percentage = "Off",
                               ever = "Off",
                               diff = "Off")

# Exports comp_table to a CSV file
# write.csv(comp_table_diff, "eligibility_table_diff.csv", row.names = TRUE)
style_compare_table(comp_hm, diff = "Off", dataname = "SCCS")
```


This chunk renders and exports the heatmap view of `comp_hm`, using 40 bins and relative-difference coloring. The HTML output is then screenshotted to PNG for reuse outside the Rmd.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
comp_hm_40 <- style_compare_heatmap(comp_hm, difference="Off", save_as_html = "comp_hm_SCCS_40.html", rel_diff = TRUE, include_legend = FALSE, bins = 40)

comp_hm_40

webshot(
  url        = "comp_hm_SCCS_40.html",
  file       = "comp_hm_SCCS_40.png",
  selector = "table",        # ← clip to the first <table> element
  vwidth     = 3600,         # or whatever width in pixels you need
  vheight    = 2000,          # adjust so that the entire table + legend fits
  zoom     = 5               # (we’ll adjust zoom below)
)
```


This chunk repeats the no-difference comparison on the ever-smoker scale so the same quantities can be viewed with an ever-smoker denominator before generating the alternate heatmap export.

```{r}
# Repeat the no-difference comparison on the ever-smoker scale for the alternate heatmap export.
comp_hm_ev <- compare_Data_SHG(elig_sccs_full_combined,
                               elig_sccs_wh_combined,
                               elig_sccs_bl_combined,
                               smpl_rslts_eligible_cl, 
                               smpl_rslts_elig_wh_cl, 
                               smpl_rslts_elig_bl_cl,
                               total = "Off",
                               comparison = "All", 
                               percentage = "On",
                               ever = "On",
                               diff = "Off")

# Exports comp_table to a CSV file
# write.csv(comp_table_diff, "eligibility_table_diff.csv", row.names = TRUE)
style_compare_table(comp_hm_ev, diff = "Off", dataname = "SCCS")
```


This chunk renders and exports the ever-smoker-normalized heatmap view of `comp_hm_ev`, again saving both the HTML version and a PNG snapshot.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
comp_hmev_40 <- style_compare_heatmap(comp_hm_ev, difference="Off", save_as_html = "comp_hmev_SCCS_40.html", rel_diff = TRUE, include_legend = TRUE, bins = 40)

comp_hmev_40

webshot(
  url        = "comp_hmev_SCCS_40.html",
  file       = "comp_hmev_SCCS_40.png",
  selector = "table",        # ← clip to the first <table> element
  vwidth     = 3600,         # or whatever width in pixels you need
  vheight    = 2000,          # adjust so that the entire table + legend fits
  zoom     = 5               # (we’ll adjust zoom below)
)
```


This line saves the entire R workspace, including all current objects (data, functions, results), to a file named backup_WS.RData. This allows for easy restoration of the session later without needing to re-run previous code.

```{r}
save.image("backup_WS.RData")
```


# Forest Plot
This chunk generates the main forest plot summarizing SHG percent error relative to SCCS across the eligibility and smoking-history measures. The resulting figure is printed and exported as a PNG.

```{r}
# Forest plot export for the main SCCS-vs-SHG eligibility comparison.
p <- plot_deviance_forest(comp_hm, xlim = 2.5)
print(p)

# Save the plot as PNG with 800x1000 pixel dimensions
ggsave(
  filename = "SCCS_forest.PNG", 
  plot = p,
  width = 800 / 100,  # Convert pixels to inches (assuming 300 dpi)
  height = 800 / 100, 
  dpi = 300
)
```


This line saves the entire R workspace, including all current objects (data, functions, results), to a file named backup_WS.RData. This allows for easy restoration of the session later without needing to re-run previous code.

```{r}
save.image("backup_WS.RData")
```


# Scatter Plots
This chunk generates and saves a 2x2 grid of scatter plots comparing smoking duration and cigarettes per day between SCCS data and SHG simulations. The compare_scatter_facet() function creates the plots for four subpopulations (Black Female, Black Male, White Female, and White Male), which are stored in SYvCPD. The resulting figure is then saved as a high-resolution PNG file named "SYvCPD_SCCS.PNG" with specified dimensions.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Raw scatter version keeps individual-level jittered points.
SYvCPD <- compare_scatter_facet(SCCS_sh_cleaned, sim_comp_rslt, threshold = "Both", arrowlabel = TRUE, bubble = FALSE)

SYvCPD

# Save the plot as PNG with 800x1000 pixel dimensions
ggsave(
  filename = "SYvCPD_SCCS.PNG", 
  plot = SYvCPD,
  width = 800 / 100,  # Convert pixels to inches (assuming 300 dpi)
  height = 800 / 100, 
  dpi = 300
)
```


This chunk creates a single binned scatter/bubble plot outside the 2x2 facet layout. It functions as a quick diagnostic panel for tuning the categorical scatter settings before regenerating the faceted figure.

```{r}
# One-off single-panel check used while tuning the binned scatter/bubble display.
g <- compare_scatter_SCCS_SHG_binned(
    SCCS_sh_cleaned, sim_comp_rslt,
    plot_title      = "test",
    x_title         = NULL,
    y_title         = "Smoking Duration (Years)",
    legend_on       = TRUE,
    seed            = 42,
    coverage        = 1,
    threshold       = "Both",
    shg_categorical = TRUE,
    arrowlabel      = TRUE
  )

g
```


This chunk generates and saves the faceted bubble version of the SCCS-versus-SHG scatter comparison, where each point represents a binned share rather than an individual-level jittered observation.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Bubble version uses binned counts/shares instead of individual-level jitter.
SYvCPD_cat <- compare_scatter_facet(SCCS_sh_cleaned, sim_comp_rslt, 
                                    threshold = "Both", bubble = TRUE, 
                                    arrowlabel = TRUE,
                                    facet_title = NULL)

SYvCPD_cat

# Save the plot as PNG with 800x1000 pixel dimensions
ggsave(
  filename = "SYvCPD_SCCS_cat.PNG", 
  plot = SYvCPD_cat,
  width = 800 / 100,  # Convert pixels to inches (assuming 300 dpi)
  height = 800 / 100, 
  dpi = 300
)
```


This line saves the entire R workspace, including all current objects (data, functions, results), to a file named backup_WS.RData. This allows for easy restoration of the session later without needing to re-run previous code.

```{r}
save.image("backup_WS.RData")
```


# Prevalence Plots
This chunk prepares data for prevalence plots by summarizing SCCS data into prevalence metrics for different populations using the dat2age function. Specifically:
	1.	dat_bl_sum: Summarizes data for the black population (RaceAnalysis == 2).
	2.	dat_wh_sum: Summarizes data for the white population (RaceAnalysis == 1).
	3.	dat_full_sum: Summarizes data for the full SCCS population without filtering by race.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Prevalence Data Preparation
# dat2age processes SCCS data into age-specific prevalence summaries used by `prvlnc_plot`.

# Create summarized prevalence data for the black population
dat_bl_sum <- dat2age(
  SCCS_sh[which(SCCS_sh$RaceAnalysis == 2), ]  # Filter SCCS data for RaceAnalysis == 2 (black population)
)

# Create summarized prevalence data for the white population
dat_wh_sum <- dat2age(
  SCCS_sh[which(SCCS_sh$RaceAnalysis == 1), ]  # Filter SCCS data for RaceAnalysis == 1 (white population)
)

# Create summarized prevalence data for the entire population
dat_full_sum <- dat2age(
  SCCS_sh  # Use the full SCCS dataset without any race filtering
)
```


This chunk generates a combined visualization of smoking prevalence for three populations: full population, white population, and black population. Each prevalence plot compares SHG-simulated smoking data to observed SCCS data. The resulting figure includes:
	1.	Prevalence of Ever Smokers and Prevalence of Current Smokers for each population.
	2.	A title for the figure (“Smoking Prevalences”).
	3.	A single, shared legend placed at the bottom of the combined figure.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Smoking Prevalence Plots

# Create the prevalence plot for the full population
prvlnc_full_plot <- prvlnc_plot(
  prvlnc_SHG_smpld_clean,    # Simulated SHG prevalence data for the full population
  dat_full_sum,        # Observed prevalence data for the full population
  subtitle = "On",     # Subtitle toggle (will show plot-specific titles)
  population = "Full Population",  # Label for the population
  CI = 99,             # Confidence interval for SHG prevalence data
  legend = "Off"       # Turn off legend for this plot
)

# Create the prevalence plot for the white population
prvlnc_wh_plot <- prvlnc_plot(
  prvlnc_SHG_wh_smpld_clean, # Simulated SHG prevalence data for the white population
  dat_wh_sum,          # Observed prevalence data for the white population
  population = "White Population", # Label for the population
  CI = 99,             # Confidence interval for SHG prevalence data
  legend = "Off"       # Turn off legend for this plot
)

# Create the prevalence plot for the black population
prvlnc_bl_plot <- prvlnc_plot(
  prvlnc_SHG_bl_smpld_clean, # Simulated SHG prevalence data for the black population
  dat_bl_sum,          # Observed prevalence data for the black population
  population = "Black Population", # Label for the population
  CI = 99,             # Confidence interval for SHG prevalence data
  legend = "On"        # Turn on legend for this plot (shared across all)
)

# Create a figure title as a standalone ggplot object
title_prvlnc <- ggplot() +
  ggtitle("Smoking Prevalences") +  # Main title for the combined plot
  theme_void() +                    # No axes or gridlines
  theme(
    plot.title = element_text(hjust = 0.5, size = 13, face = "bold")  # Center and style the title
  )

# Combine all prevalence plots into a single figure
prvlnc_fig <- (
  title_prvlnc /                 # Add the main title on top
  prvlnc_full_plot /             # Add the full population prevalence plot
  prvlnc_wh_plot /               # Add the white population prevalence plot
  prvlnc_bl_plot                 # Add the black population prevalence plot
) +
  plot_layout(
    heights = c(0, 1, 1, 1),     # Allocate zero height for the title and equal heights for plots
    guides = "collect"           # Collect all legends into a single shared legend
  ) +
  plot_annotation(
    theme = theme(
      legend.position = "bottom",    # Place the legend at the bottom
      legend.box = "horizontal",     # Arrange legend items in a horizontal box
      legend.title = element_blank() # Remove the legend title
    )
  )

# Render the combined prevalence plot
prvlnc_fig
```


This line saves the entire R workspace, including all current objects (data, functions, results), to a file named backup_WS.RData. This allows for easy restoration of the session later without needing to re-run previous code.

```{r}
save.image("backup_WS.RData")
```


# Median Tables
This section prepares helper functions for the median/IQR summary tables. The intended workflow is to compute empirical SCCS quantiles directly from each observed subgroup, then create one large subgroup-matched SHG sample by inflating that subgroup's birth-year/sex target by a factor of 10 and sampling once from the large SHG pool. The resulting observed and simulated summaries can then be stacked with a percent-difference row for the medians.

```{r}
make_byr_target <- function(df, coverage = 10, collection_year = data_collection_year) {
  if (!"YOB" %in% names(df)) {
    df <- df %>%
      mutate(YOB = collection_year - Enrollment_Age)
  }

  byr <- df %>%
    group_by(YOB) %>%
    summarise(
      Females = sum(Sex == "F", na.rm = TRUE),
      Males   = sum(Sex == "M", na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(YOB)

  sufftrgt(byr, coverage = coverage)
}


summarize_smoking_quantiles <- function(df, dataset_type = c("SCCS", "SHG")) {
  dataset_type <- match.arg(dataset_type)

  if (dataset_type == "SCCS") {
    ever_col <- "EverSmoker"
    pack_col <- "PackYears"
    dur_col  <- "SmokingYears"
  } else {
    ever_col <- "Eversmoker"
    pack_col <- "Pack_Years"
    dur_col  <- "Smoking_Years"
  }

  df_ever <- df %>%
    filter(.data[[ever_col]] == 1)

  tibble(
    CigsPerDay_Median = median(df_ever$CigsPerDay, na.rm = TRUE),
    CigsPerDay_IQR_Low = quantile(df_ever$CigsPerDay, 0.25, na.rm = TRUE),
    CigsPerDay_IQR_High = quantile(df_ever$CigsPerDay, 0.75, na.rm = TRUE),
    PackYears_Median = median(df_ever[[pack_col]], na.rm = TRUE),
    PackYears_IQR_Low = quantile(df_ever[[pack_col]], 0.25, na.rm = TRUE),
    PackYears_IQR_High = quantile(df_ever[[pack_col]], 0.75, na.rm = TRUE),
    SmokingYears_Median = median(df_ever[[dur_col]], na.rm = TRUE),
    SmokingYears_IQR_Low = quantile(df_ever[[dur_col]], 0.25, na.rm = TRUE),
    SmokingYears_IQR_High = quantile(df_ever[[dur_col]], 0.75, na.rm = TRUE)
  )
}


make_median_diff_row <- function(observed_row, simulated_row) {
  tibble(
    CigsPerDay_Median = 100 * (simulated_row$CigsPerDay_Median - observed_row$CigsPerDay_Median) / observed_row$CigsPerDay_Median,
    CigsPerDay_IQR_Low = NA_real_,
    CigsPerDay_IQR_High = NA_real_,
    PackYears_Median = 100 * (simulated_row$PackYears_Median - observed_row$PackYears_Median) / observed_row$PackYears_Median,
    PackYears_IQR_Low = NA_real_,
    PackYears_IQR_High = NA_real_,
    SmokingYears_Median = 100 * (simulated_row$SmokingYears_Median - observed_row$SmokingYears_Median) / observed_row$SmokingYears_Median,
    SmokingYears_IQR_Low = NA_real_,
    SmokingYears_IQR_High = NA_real_
  )
}


build_median_table <- function(observed_df, shg_pool, coverage = 10, seed = 42) {
  byr_target <- make_byr_target(observed_df, coverage = coverage)
  simulated_sample <- sample_SHG(shg_pool, byr_target, seed = seed)

  observed_row <- summarize_smoking_quantiles(observed_df, dataset_type = "SCCS")
  simulated_row <- summarize_smoking_quantiles(simulated_sample, dataset_type = "SHG")
  diff_row <- make_median_diff_row(observed_row, simulated_row)

  bind_rows(
    Observed = observed_row,
    Simulated = simulated_row,
    `% Difference` = diff_row,
    .id = "Source"
  )
}
```


This chunk applies the median-table workflow to the six SCCS subpopulations used elsewhere in the analysis, then stacks the results into one long export table. Each population contributes three rows: Observed, Simulated, and % Difference.

```{r}
median_table_inputs <- list(
  Blacks = SCCS_sh_cleaned %>% filter(RaceAnalysis == 2),
  Whites = SCCS_sh_cleaned %>% filter(RaceAnalysis == 1),
  `Black Females` = SCCS_sh_cleaned %>% filter(RaceAnalysis == 2, Sex == "F"),
  `Black Males` = SCCS_sh_cleaned %>% filter(RaceAnalysis == 2, Sex == "M"),
  `White Females` = SCCS_sh_cleaned %>% filter(RaceAnalysis == 1, Sex == "F"),
  `White Males` = SCCS_sh_cleaned %>% filter(RaceAnalysis == 1, Sex == "M")
)

median_tables_long <- purrr::imap_dfr(
  median_table_inputs,
  ~ build_median_table(.x, sim_comp_rslt, coverage = 10, seed = 42) %>%
    mutate(Population = .y, .before = 1)
)

median_tables_long

write.csv(median_tables_long, "SCCS_median_tables_long.csv", row.names = FALSE)
```


# Threshold Tables
This section builds two long threshold tables from the existing comparison-table objects already in the workspace. One table reports percentages relative to ever-smokers, and the other reports percentages relative to the full subgroup population. In each case, the simulated row keeps the SHG mean and confidence interval, while the Difference row reports the simulated mean minus the observed value.

```{r}
parse_shg_summary <- function(x) {
  nums <- stringr::str_extract_all(x, "-?[0-9.]+")[[1]]
  tibble(
    Value = as.numeric(nums[1]),
    Lower_CI = as.numeric(nums[2]),
    Upper_CI = as.numeric(nums[3])
  )
}


build_threshold_table_from_compare <- function(comp_table) {
  threshold_rows <- c("PY20", "PY30", "SY20", "SY30")
  populations <- c("Black", "White", "Black Female", "Black Male", "White Female", "White Male")
  population_labels <- c(
    Black = "Blacks",
    White = "Whites",
    `Black Female` = "Black Females",
    `Black Male` = "Black Males",
    `White Female` = "White Females",
    `White Male` = "White Males"
  )

  purrr::map_dfr(populations, function(pop) {
    observed_vals <- purrr::map_dbl(threshold_rows, ~ as.numeric(comp_table[.x, paste0(pop, "_Data")]))
    simulated_vals <- purrr::map(
      threshold_rows,
      ~ parse_shg_summary(as.character(comp_table[.x, paste0(pop, "_SHG")]))
    )
    simulated_means <- purrr::map_dbl(simulated_vals, ~ .x$Value)
    simulated_lowers <- purrr::map_dbl(simulated_vals, ~ .x$Lower_CI)
    simulated_uppers <- purrr::map_dbl(simulated_vals, ~ .x$Upper_CI)
    diff_vals <- simulated_means - observed_vals

    bind_rows(
      tibble(
        Population = population_labels[[pop]],
        Source = "Observed",
        PY20_Value = observed_vals[1],
        PY20_Lower_CI = NA_real_,
        PY20_Upper_CI = NA_real_,
        PY30_Value = observed_vals[2],
        PY30_Lower_CI = NA_real_,
        PY30_Upper_CI = NA_real_,
        SY20_Value = observed_vals[3],
        SY20_Lower_CI = NA_real_,
        SY20_Upper_CI = NA_real_,
        SY30_Value = observed_vals[4],
        SY30_Lower_CI = NA_real_,
        SY30_Upper_CI = NA_real_
      ),
      tibble(
        Population = population_labels[[pop]],
        Source = "Simulated",
        PY20_Value = simulated_means[1],
        PY20_Lower_CI = simulated_lowers[1],
        PY20_Upper_CI = simulated_uppers[1],
        PY30_Value = simulated_means[2],
        PY30_Lower_CI = simulated_lowers[2],
        PY30_Upper_CI = simulated_uppers[2],
        SY20_Value = simulated_means[3],
        SY20_Lower_CI = simulated_lowers[3],
        SY20_Upper_CI = simulated_uppers[3],
        SY30_Value = simulated_means[4],
        SY30_Lower_CI = simulated_lowers[4],
        SY30_Upper_CI = simulated_uppers[4]
      ),
      tibble(
        Population = population_labels[[pop]],
        Source = "Difference",
        PY20_Value = diff_vals[1],
        PY20_Lower_CI = NA_real_,
        PY20_Upper_CI = NA_real_,
        PY30_Value = diff_vals[2],
        PY30_Lower_CI = NA_real_,
        PY30_Upper_CI = NA_real_,
        SY20_Value = diff_vals[3],
        SY20_Lower_CI = NA_real_,
        SY20_Upper_CI = NA_real_,
        SY30_Value = diff_vals[4],
        SY30_Lower_CI = NA_real_,
        SY30_Upper_CI = NA_real_
      )
    )
  })
}
```


This chunk assembles the threshold table where all four measures are expressed as percentages of ever-smokers. The simulated entries retain their SHG confidence intervals, and the Difference row records the simulated mean minus the observed value.

```{r}
threshold_table_ever_long <- build_threshold_table_from_compare(comp_table_ev)

threshold_table_ever_long

write.csv(threshold_table_ever_long, "SCCS_threshold_table_ever_long.csv", row.names = FALSE)
```


This chunk assembles the threshold table where all four measures are expressed as percentages of the total subgroup population. As above, the simulated row retains the SHG confidence intervals and the Difference row reports the simulated mean minus the observed value.

```{r}
threshold_table_total_long <- build_threshold_table_from_compare(comp_table_cl)

threshold_table_total_long

write.csv(threshold_table_total_long, "SCCS_threshold_table_total_long.csv", row.names = FALSE)
```


# Probability Tables
This section summarizes directional distributional differences between observed SCCS data and subgroup-matched SHG samples. For each smoking-history measure, it reports the probability that a randomly drawn SHG value exceeds a randomly drawn SCCS value, along with the centered transformation `2P - 1`, where 0 indicates no directional shift.

```{r}
prob_superiority <- function(sim_values, obs_values) {
  sim_values <- sim_values[!is.na(sim_values)]
  obs_values <- obs_values[!is.na(obs_values)]

  if (length(sim_values) == 0 || length(obs_values) == 0) {
    return(NA_real_)
  }

  combined <- c(sim_values, obs_values)
  ranks <- rank(combined, ties.method = "average")

  n_sim <- as.numeric(length(sim_values))
  n_obs <- as.numeric(length(obs_values))
  rank_sum_sim <- sum(ranks[seq_len(n_sim)])

  (rank_sum_sim - n_sim * (n_sim + 1) / 2) / (n_sim * n_obs)
}


build_probability_row <- function(observed_df, shg_pool, coverage = 10, seed = 42) {
  simulated_sample <- sample_SHG(shg_pool, make_byr_target(observed_df, coverage = coverage), seed = seed)

  observed_ever <- observed_df %>% filter(EverSmoker == 1)
  simulated_ever <- simulated_sample %>% filter(Eversmoker == 1)

  p_cigs <- prob_superiority(simulated_ever$CigsPerDay, observed_ever$CigsPerDay)
  p_pack <- prob_superiority(simulated_ever$Pack_Years, observed_ever$PackYears)
  p_smoke <- prob_superiority(simulated_ever$Smoking_Years, observed_ever$SmokingYears)

  tibble(
    CigsPerDay_P = p_cigs,
    CigsPerDay_2Pminus1 = 2 * p_cigs - 1,
    PackYears_P = p_pack,
    PackYears_2Pminus1 = 2 * p_pack - 1,
    SmokingYears_P = p_smoke,
    SmokingYears_2Pminus1 = 2 * p_smoke - 1
  )
}
```


This chunk applies the probability-of-superiority calculation to the six SCCS subpopulations used elsewhere in the analysis. Each row summarizes the directional shift between the observed subgroup distribution and a single large subgroup-matched SHG sample.

```{r}
probability_tables_long <- purrr::imap_dfr(
  median_table_inputs,
  ~ build_probability_row(.x, sim_comp_rslt, coverage = 10, seed = 42) %>%
    mutate(Population = .y, .before = 1)
)

probability_tables_long

write.csv(probability_tables_long, "SCCS_probability_tables_long.csv", row.names = FALSE)
```


This line saves the entire R workspace, including all current objects (data, functions, results), to a file named backup_WS.RData. This allows for easy restoration of the session later without needing to re-run previous code.

```{r}
save.image("backup_WS.RData")
```
